---
title: "Does physical inactivity level and access to healthy food affect obesity rate at the county level?"
author: Helen Peng, Anvith Thumma, and Jordan Faustin
format: 
  revealjs:
    theme: simple
    slide-number: true
    smaller: true
editor: visual
---

## Why This Question is Important

-   **Public Health Crisis:** Obesity is a major contributor to chronic diseases (diabetes, heart disease, cancer) and rising healthcare costs.

-   **Disparities in Obesity:** Obesity rates vary widely by region, income, and race/ethnicity. Identifying structural barriers (like food access and inactivity) can inform equitable policies.

-   **Rise in Weight Loss Drugs:** Medications like Ozempic and Wegovy are gaining popularity as quick fixes, but they’re expensive and don’t address root causes like environment or behavior.

## Data Overview

-   **Observations:** Data at the county level across all 50 U.S. states in 2025

-   **Variables of Interest:**

    -   \% Adults with Obesity

    -   \% Physically Inactive

    -   Food Environment Index (1-10) - Higher is Better

    -   State and County Identifiers

-   **Handling Missing Data:**

    -   If data on Food Environment Index or Physical Inactivity was missing for a county we used the state average.

    -   Seven counties that did not have data for obesity rate, our response variable and so, we dropped those counties to do a complete case analysis.

```{r setup}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center",
               echo=FALSE, tidy=TRUE)
```

```{r cleaning}
library(readxl)
library(tidyverse)
theme_set(theme_bw())

data_2025 <- read_excel("C:/Users/helen/OneDrive/Documents/GitHub/obesity-capstone/Peng/2025_data.xlsx", sheet = "Additional Measure Data", skip = 1)

food_index_2025 <- read_excel("C:/Users/helen/OneDrive/Documents/GitHub/obesity-capstone/Peng/2025_data.xlsx", sheet = "Select Measure Data", skip = 1)

data_2025 <- data_2025 |> 
  janitor::clean_names() 

food_index_2025 <- food_index_2025 |> 
  janitor::clean_names() 

data_2025 <- merge(data_2025, food_index_2025, by = "fips") 
data_2025 <- data_2025[, !grepl("percent_ci", names(data_2025))]
data_2025 <- data_2025[, !grepl("national_z", names(data_2025))]
data_2025 <- data_2025[, !grepl("injury_death", names(data_2025))]
data_2025 <- data_2025[, !grepl("percentile", names(data_2025))]
data_2025 <- data_2025[, !grepl("_white", names(data_2025))]
data_2025 <- data_2025[, !grepl("_hispanic", names(data_2025))]
data_2025 <- data_2025[, !grepl("_native", names(data_2025))]
data_2025 <- data_2025[, !grepl("_aian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_asian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_black", names(data_2025))]
data_2025 <- data_2025[, !grepl("dentist", names(data_2025))]
data_2025 <- data_2025[, !grepl("unreliable", names(data_2025))]
data_2025 <- data_2025[, !grepl("deaths", names(data_2025))]

drop_cols <- c("child_mortality_rate","hiv_prevalence_rate", " suicide_rate_age_adjusted", "crude_rate","teen_birth_rate","chlamydia_rate","number_chlamydia_cases", "drug_overdose_mortality_rate","other_primary_care_provider_rate", "other_primary_care_provider_ratio","percent_census_participation", "average_grade_performance_252", " average_grade_performance_246","cohort_size","gender_pay_gap","segregation_index_287","homicide_rate","motor_vehicle_mortality_rate", "percent_disconnected_youth","firearm_fatalities_rate","percent_lacking_support", "percent_low_birth_weight","number_primary_care_physicians","primary_care_physicians_ratio","primary_care_physicians_rate","number_mental_health_providers", "mental_health_provider_rate", "mental_health_provider_ratio", "visits_per_service_area_population", "presence_of_water_violation","average_daily_pm2_5", "state.y","county.y", "percent_with_annual_mammogram", "preventable_hospitalization_rate", "years_of_potential_life_lost_rate","number_firearm_fatalities","child_care_centers_per_1_000_children", "number_firearm_fatalities", "number_child_care_centers","spending_per_pupil","segregation_index_258","high_school_graduation_rate","average_grade_performance_246","disaster_declarations","weeks_in_moderate_or_greater_drought", "life_expectancy", "severe_housing_cost_burden", "percent_adults_with_diabetes","suicide_rate_age_adjusted","percent_fair_or_poor_health", "percent_voter_turnout", "percent_with_disability", "age_adjusted_death_rate", "percent_drive_alone_to_work", "income_ratio", "mens_median_earnings", "womens_median_earnings", "infant_mortality_rate","percent_uninsured_children","percent_uninsured_adults","percent_with_access_to_exercise_opportunities")

data_2025 <- data_2025[,!(names(data_2025) %in% drop_cols)]

data_2025 <- data_2025  |> 
  rename(
    county = county.x,
    state = state.x
  )

old_ct_fips <- c("09001", "09003", "09005", "09007", "09009", "09011", "09013", "09015")  

statewide_food_index <- data_2025 |> 
  filter(is.na(county) & !is.na(food_environment_index)) |> 
  dplyr::select(state, food_environment_index) |> 
  rename(statewide_food_environment_index = food_environment_index)

data_2025 <- data_2025 |> 
  filter(!(fips %in% old_ct_fips)) |> 
  left_join(statewide_food_index, by = "state") |> 
  mutate(
    food_environment_index = if_else(
      is.na(food_environment_index) & !is.na(county),
      statewide_food_environment_index,
      food_environment_index),
      days_above_90f = if_else(days_above_90f > 365, 365, days_above_90f)) |> 
  dplyr::select(-statewide_food_environment_index) |> 
  filter(!is.na(percent_adults_with_obesity))
```

## As physical inactivity increases obesity rates increases and as access to healthy food increases obesity rates decreases.

```{r scatterplot}
library(patchwork)

physical_obesity_plot <- data_2025 |>
  ggplot(aes(x = percent_physically_inactive, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "% Physically Inactive",
    y = "% Adults with Obesity"
  ) +
  theme_minimal()

food_obesity_plot <- data_2025 |>
  ggplot(aes(x = food_environment_index, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "Food Environment Index (0 = worst, 10 = best)",
    y = "% Adults with Obesity"
  ) +
  theme_minimal()

physical_obesity_plot + food_obesity_plot 
```

## Obesity rates are consistently higher in the Midwest and the South compared to the West Coast and the Northeast.

```{r obesity_map}
library(usmap)

q_brks <- floor(quantile(data_2025$percent_adults_with_obesity, probs = seq(0, 1, length.out = 5), na.rm = TRUE))

plot_usmap( data = data_2025, values = "percent_adults_with_obesity") +
  scale_fill_gradient(low = "lightyellow", high = "darkred",
    breaks = q_brks,
    labels = round(q_brks, 1),
    name = "Obesity Rate (%)",
    na.value = "grey90"
  ) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "bold")
  )
```

## Surprisingly, the Food Environment Index does not rank among the top predictors of county-level obesity rates.

```{r vip}

# CART-algorthim regression tree 10 fold CV
library(caret)
library(vip)

set.seed(10)
data_2025 <- data_2025 |> na.omit()

tree <- train(percent_adults_with_obesity ~ . - fips - state - county,
              method = "rpart", tuneLength = 20,
              trControl = trainControl(method = "cv", number = 10),
              data = data_2025)

top_vi <- head(vi(tree), 15)

# pretty_labels <- c(
#   "percent_physically_inactive" = "Physically Inactive",
#   "median_household_income" = "Median Household Income",
#   "percent_adults_reporting_currently_smoking" = "Smoking",
#   "percent_excessive_drinking" = "Excessive Drinking",
#   "percent_below_18_years_of_age" = "Population Below 18 Years Old",
#   "percent_insufficient_sleep" = "Insufficient Sleep",
#   "school_funding_adequacy" = "School Funding Adequacy",
#   "percent_households_with_severe_cost_burden" = "Severe Cost Burden",
#   "percent_uninsured_adults" = "Uninsured Adults",
#   "traffic_volume" = "Traffic Volume",
#   "food_environment_index" = "Food Environment Index"
# )
# 
# top_vi$Variable <- ifelse(
#   top_vi$Variable %in% names(pretty_labels),
#   pretty_labels[top_vi$Variable],
#   top_vi$Variable
# )

vip(top_vi, aesthetics = list(fill = "steelblue")) +
  labs(y = "Variable Importance (Total Reduction in RSS)")

```

## Linear Regression Model

```{r bivariate_model}
library(MASS)

data_2025 <- data_2025 |> 
  mutate(
    obesity_prop = percent_adults_with_obesity / 100,
    logit_obesity = log(obesity_prop / (1 - obesity_prop))
  )

lm_simple <- lm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
    data = data_2025)

summary(lm_simple)
# plot(lm_simple) -> Moved to appendix

huber_model <- rlm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
    data = data_2025)

summary(huber_model)
# Need to justify this over lm_simple
```
```{r multivariate_model}

```

## Model Comparsion


## Recommendations

-   Prioritize the young population by targeting environments that shape long-term health behaviors early in life.

-   Enhance sidewalk infrastructure across all clusters to promote walkability, encourage physical activity, and support safe routes for children and families.

-   Invest in school funding to strengthen the local food environment index and promote healthy habits—such as walking to school and accessing nutritious, affordable meals.


## Appendix

- Correlation Matrix
- Diagnostic Plots for Linear Regression Models
- 

