---
title: "Mapping Obesity: A County-Level Analysis of Lifestyle and Food Environment Factors"
subtitle: "SURE 2025"
author:
  - "<strong>Helen Peng</strong><br><small>Carnegie Mellon University</small>"
  - "<strong>Jordan Faustin</strong><br><small>Howard University</small>"
  - "<strong>Anvith Thumma</strong><br><small>Arizona State University</small>"
format: 
  revealjs: 
    embed-resources: true
    theme: simple
    slide-number: true
    smaller: true
    include-after-body: |
      <style>
        .reveal .slides section.title-slide {
          text-align: center;
        }
        .reveal .slides section.title-slide .author p {
          margin: 1em 0;
          line-height: 1.5;
        }
      </style>
editor: visual
---

## Significance

-   **Public Health Implications:** <br> Obesity is a major contributor to rising healthcare costs and chronic diseases such as diabetes, heart disease, and cancer.

-   **Social Factors of Health:** <br> Physical Inactivity and food access are modifiable factors linked to obesity. In turn, studying their impact can highlight how our environment and infrastructure, like our parks, grocery stores, and schools, can shape our health outcomes.

-   **Policy and Community Planning:** <br> New information revealed through findings can help to make informed policies for urban planning in under-served areas and school/workplace wellness programs.

## Does physical inactivity level and access to healthy food affect obesity rate at the county level? {.center}

## Agenda

-   **Data Overview**
-   **Initial Analysis**
-   **Understanding Direction**
-   **Modeling and Comparison**
-   **Discussion**
-   **Recommendations**
-   **Appendix**

## Data Overview

-   **Observations:**

    -   Data at the county level across all 50 U.S. states in 2025

-   **Variables of Interest:**

    -   \% Adults with Obesity

    -   \% Physically Inactive

    -   (FEI) Food Environment Index (1-10)

    -   \% Limited Access to Healthy Foods

    -   State and County Identifiers

-   **Handling Missing Data:**

    -   If data on Food Environment Index or Physical Inactivity was missing for a county we used the state average.

    -   Seven counties that did not have data for obesity rate, our response variable and so, we dropped those counties to do a complete case analysis.

```{r setup}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center",
               echo=FALSE, tidy=TRUE)
```

```{r cleaning}
library(readxl)
library(tidyverse)
theme_set(theme_bw())

data_2025 <- read_excel("C:/Users/jorda/Desktop/Capstone Obesity/2025_CHRD.xlsx", sheet = "Additional Measure Data", skip = 1)

food_index_2025 <- read_excel("C:/Users/jorda/Desktop/Capstone Obesity/2025_CHRD.xlsx", sheet = "Select Measure Data", skip = 1)

data_2025 <- data_2025 |> 
  janitor::clean_names() 

food_index_2025 <- food_index_2025 |> 
  janitor::clean_names() 

data_2025 <- merge(data_2025, food_index_2025, by = "fips") 
data_2025 <- data_2025[, !grepl("percent_ci", names(data_2025))]
data_2025 <- data_2025[, !grepl("national_z", names(data_2025))]
data_2025 <- data_2025[, !grepl("injury_death", names(data_2025))]
data_2025 <- data_2025[, !grepl("percentile", names(data_2025))]
data_2025 <- data_2025[, !grepl("_white", names(data_2025))]
data_2025 <- data_2025[, !grepl("_hispanic", names(data_2025))]
data_2025 <- data_2025[, !grepl("_native", names(data_2025))]
data_2025 <- data_2025[, !grepl("_aian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_asian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_black", names(data_2025))]
data_2025 <- data_2025[, !grepl("dentist", names(data_2025))]
data_2025 <- data_2025[, !grepl("unreliable", names(data_2025))]
data_2025 <- data_2025[, !grepl("deaths", names(data_2025))]

drop_cols <- c("child_mortality_rate","hiv_prevalence_rate", " suicide_rate_age_adjusted", "crude_rate","teen_birth_rate","chlamydia_rate","number_chlamydia_cases", "drug_overdose_mortality_rate","other_primary_care_provider_rate", "other_primary_care_provider_ratio","percent_census_participation", "average_grade_performance_252", " average_grade_performance_246","cohort_size","gender_pay_gap","segregation_index_287","homicide_rate","motor_vehicle_mortality_rate", "percent_disconnected_youth","firearm_fatalities_rate","percent_lacking_support", "percent_low_birth_weight","number_primary_care_physicians","primary_care_physicians_ratio","primary_care_physicians_rate","number_mental_health_providers", "mental_health_provider_rate", "mental_health_provider_ratio", "visits_per_service_area_population", "presence_of_water_violation","average_daily_pm2_5", "state.y","county.y", "percent_with_annual_mammogram", "preventable_hospitalization_rate", "years_of_potential_life_lost_rate","number_firearm_fatalities","child_care_centers_per_1_000_children", "number_firearm_fatalities", "number_child_care_centers","spending_per_pupil","segregation_index_258","high_school_graduation_rate","average_grade_performance_246","disaster_declarations","weeks_in_moderate_or_greater_drought", "life_expectancy", "severe_housing_cost_burden", "percent_adults_with_diabetes","suicide_rate_age_adjusted","percent_fair_or_poor_health", "percent_voter_turnout", "percent_with_disability", "age_adjusted_death_rate", "percent_drive_alone_to_work", "income_ratio", "mens_median_earnings", "womens_median_earnings", "infant_mortality_rate","percent_uninsured_children","percent_uninsured_adults","percent_with_access_to_exercise_opportunities")

data_2025 <- data_2025[,!(names(data_2025) %in% drop_cols)]

data_2025 <- data_2025  |> 
  rename(
    county = county.x,
    state = state.x
  )

old_ct_fips <- c("09001", "09003", "09005", "09007", "09009", "09011", "09013", "09015")  

statewide_food_index <- data_2025 |> 
  filter(is.na(county) & !is.na(food_environment_index)) |> 
  dplyr::select(state, food_environment_index) |> 
  rename(statewide_food_environment_index = food_environment_index)

data_2025 <- data_2025 |> 
  filter(!(fips %in% old_ct_fips)) |> 
  left_join(statewide_food_index, by = "state") |> 
  mutate(
    food_environment_index = if_else(
      is.na(food_environment_index) & !is.na(county),
      statewide_food_environment_index,
      food_environment_index),
      days_above_90f = if_else(days_above_90f > 365, 365, days_above_90f)) |> 
  dplyr::select(-statewide_food_environment_index) |> 
  filter(!is.na(percent_adults_with_obesity))
```

# Mapping Obesity: A County-Level Analysis of Lifestyle and Food Environment Factors
# EDA Analysis Presentation R Code
```{r}
# Load required libraries
library(usmap)
library(viridis)
library(ggplot2)
library(dplyr)
library(caret)
library(rpart)
library(vip)
library(corrplot)
library(mgcv)
library(MASS)
library(knitr)
library(kableExtra)
```
# Assuming your data is loaded as data_2025
# Handle missing data as described
# data_2025 <- your_data_loading_code_here

# ===== SLIDE 6: Obesity Map =====
```{r}
create_obesity_map <- function(data) {
  min_val <- floor(min(data$percent_adults_with_obesity, na.rm = TRUE))
  max_val <- ceiling(max(data$percent_adults_with_obesity, na.rm = TRUE))
  custom_breaks <- seq(min_val, max_val, length.out = 5)
  
  plot_usmap(data = data, values = "percent_adults_with_obesity", color = NA) +
    scale_fill_viridis_c(
      option = "inferno",
      direction = -1,
      limits = c(min_val, max_val),
      breaks = custom_breaks,
      labels = round(custom_breaks, 1),
      name = "Obesity Rate (%)",
      na.value = "grey90"
    ) +
    theme(
      legend.position = "left",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
    ) +
    labs(title = "County-Level Obesity Rates Across the United States")
}
```

# ===== SLIDE 7: Scatterplot Analysis =====
```{r}
# Visualization 1: Physical Inactivity vs Obesity
create_inactivity_plot <- function(data) {
  ggplot(data, aes(x = percent_physically_inactive, y = percent_adults_with_obesity)) +
    geom_point(alpha = 0.6, color = "steelblue") +
    geom_smooth(method = "gam", formula = y ~ s(x), color = "red", se = TRUE) +
    labs(
      title = "Physical Inactivity vs Obesity Rate",
      x = "Percent Physically Inactive (%)",
      y = "Percent Adults with Obesity (%)"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}

# Visualization 2: Access to Healthy Food vs Obesity
create_food_access_plot <- function(data) {
  ggplot(data, aes(x = percent_limited_access_to_healthy_foods, y = percent_adults_with_obesity)) +
    geom_point(alpha = 0.6, color = "darkgreen") +
    geom_smooth(method = "gam", formula = y ~ s(x), color = "red", se = TRUE) +
    labs(
      title = "Limited Access to Healthy Foods vs Obesity Rate",
      x = "Percent with Limited Access to Healthy Foods (%)",
      y = "Percent Adults with Obesity (%)"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}
```

# ===== SLIDE 8: Food Environment Index Analysis =====
```{r}
# Visualization 3: FEI vs Obesity
create_fei_plot <- function(data) {
  ggplot(data, aes(x = food_environment_index, y = percent_adults_with_obesity)) +
    geom_point(alpha = 0.6, color = "purple") +
    geom_smooth(method = "gam", formula = y ~ s(x), color = "red", se = TRUE) +
    labs(
      title = "Food Environment Index vs Obesity Rate",
      x = "Food Environment Index (1-10 scale)",
      y = "Percent Adults with Obesity (%)",
      subtitle = "Higher FEI indicates better food environment"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5))
}
```

# ===== SLIDE 9: CART Variable Importance for FEI =====
```{r}
create_fei_importance <- function(data) {
  # Clean data
  set.seed(10)
  data_clean <- data %>% na.omit()
  
  # Train CART model
  tree_fei <- train(food_environment_index ~ . - fips - state - county,
                    method = "rpart", 
                    tuneLength = 20,
                    trControl = trainControl(method = "cv", number = 10),
                    data = data_clean)
  
  # Get top 10 predictors
  top_vi_fei <- head(vi(tree_fei), 10)
  
  # Create variable importance plot
  vip(top_vi_fei, aesthetics = list(fill = "darkgreen")) +
    labs(
      title = "Top 10 Predictors of Food Environment Index",
      y = "Variable Importance (Total Reduction in RSS)",
      x = "Variables"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}
```

# ===== SLIDE 10: Correlation Matrix =====
```{r}
create_correlation_matrix <- function(data) {
  # Select key variables for correlation
  key_vars <- data %>%
    dplyr::select(percent_adults_with_obesity, 
           percent_physically_inactive,
           food_environment_index,
           percent_limited_access_to_healthy_foods,
           percent_food_insecure,
           median_household_income,
           percent_children_in_poverty,
           percent_some_college) %>%
    na.omit()
  
  # Calculate correlation matrix
  cor_matrix <- cor(key_vars)
  
  # Create correlation plot
  corrplot(cor_matrix, 
           method = "color",
           type = "upper",
           order = "hclust",
           tl.cex = 0.8,
           tl.col = "black",
           tl.srt = 45,
           title = "Correlation Matrix: Key Health and Socioeconomic Variables",
           mar = c(0,0,2,0))
}
```

# ===== SLIDE 11: County-Level Analysis =====
```{r}
create_county_summary <- function(data) {
  # Summary statistics table
  summary_stats <- data %>%
    dplyr::select(percent_adults_with_obesity, 
           percent_physically_inactive,
           food_environment_index,
           percent_limited_access_to_healthy_foods) %>%
    na.omit() %>%
    summarise(
      across(everything(), 
             list(mean = ~round(mean(.x), 2),
                  median = ~round(median(.x), 2),
                  sd = ~round(sd(.x), 2),
                  min = ~round(min(.x), 2),
                  max = ~round(max(.x), 2)), 
             .names = "{.col}_{.fn}")
    )
  
  # Reshape for better presentation
  summary_long <- summary_stats %>%
    pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
    separate(variable, into = c("variable", "statistic"), sep = "_(?=[^_]*$)") %>%
    pivot_wider(names_from = statistic, values_from = value)
  
  return(summary_long)
}

# Create scatter plot matrix for key variables
create_scatter_matrix <- function(data) {
  key_data <- data %>%
    dplyr::select(percent_adults_with_obesity, 
           percent_physically_inactive,
           food_environment_index,
           percent_limited_access_to_healthy_foods) %>%
    na.omit()
  
  pairs(key_data, 
        main = "Scatter Plot Matrix: Key Variables",
        pch = 19, 
        cex = 0.5,
        col = alpha("steelblue", 0.6))
}
```

# ===== SLIDE 12: Regional Comparison =====
```{r}
# Add region classification function
classify_region <- function(state) {
  midwest <- c("IL", "IN", "IA", "KS", "MI", "MN", "MO", "NE", "ND", "OH", "SD", "WI")
  south <- c("AL", "AR", "DE", "FL", "GA", "KY", "LA", "MD", "MS", "NC", "OK", "SC", "TN", "TX", "VA", "WV")
  northeast <- c("CT", "ME", "MA", "NH", "NJ", "NY", "PA", "RI", "VT")
  west <- c("AK", "AZ", "CA", "CO", "HI", "ID", "MT", "NV", "NM", "OR", "UT", "WA", "WY")
  
  case_when(
    state %in% midwest ~ "Midwest",
    state %in% south ~ "South",
    state %in% northeast ~ "Northeast",
    state %in% west ~ "West",
    TRUE ~ "Other"
  )
}

create_regional_analysis <- function(data) {
  # Add region classification
  data_regional <- data %>%
    mutate(region = classify_region(state)) %>%
    filter(region != "Other") %>%
    na.omit()
  
  # Regional summary
  regional_summary <- data_regional %>%
    group_by(region) %>%
    summarise(
      n_counties = n(),
      avg_obesity = round(mean(percent_adults_with_obesity), 2),
      avg_inactivity = round(mean(percent_physically_inactive), 2),
      avg_fei = round(mean(food_environment_index), 2),
      avg_limited_food_access = round(mean(percent_limited_access_to_healthy_foods), 2),
      .groups = 'drop'
    )
  
  # Box plot comparison
  p1 <- ggplot(data_regional, aes(x = region, y = percent_adults_with_obesity, fill = region)) +
    geom_boxplot(alpha = 0.7) +
    labs(title = "Obesity Rates by Region",
         x = "Region", y = "Percent Adults with Obesity (%)") +
    theme_minimal() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold"))
  
  return(list(summary = regional_summary, plot = p1))
}
```

# ===== SLIDE 13: Huber Regression Models =====
```{r}
create_huber_models <- function(data) {
  # Clean data
  model_data <- data %>%
    dplyr::select(percent_adults_with_obesity, 
           percent_physically_inactive,
           food_environment_index,
           percent_limited_access_to_healthy_foods) %>%
    na.omit()
  
  # Check if we have enough data
  if (nrow(model_data) < 10) {
    stop("Not enough complete cases for modeling")
  }
  
  # Model 1: FEI + Physical Inactivity
  tryCatch({
    model1 <- rlm(percent_adults_with_obesity ~ food_environment_index + percent_physically_inactive,
                  data = model_data, method = "M")
  }, error = function(e) {
    # Fallback to regular linear model if Huber fails
    model1 <- lm(percent_adults_with_obesity ~ food_environment_index + percent_physically_inactive,
                 data = model_data)
  })
  
  # Model 2: Physical Inactivity + Access to Healthy Foods
  tryCatch({
    model2 <- rlm(percent_adults_with_obesity ~ percent_physically_inactive + percent_limited_access_to_healthy_foods,
                  data = model_data, method = "M")
  }, error = function(e) {
    # Fallback to regular linear model if Huber fails
    model2 <- lm(percent_adults_with_obesity ~ percent_physically_inactive + percent_limited_access_to_healthy_foods,
                 data = model_data)
  })
  
  # Create residual plots
  model_data$fitted1 <- fitted(model1)
  model_data$residuals1 <- residuals(model1)
  model_data$fitted2 <- fitted(model2)
  model_data$residuals2 <- residuals(model2)
  
  p1 <- ggplot(model_data, aes(x = fitted1, y = residuals1)) +
    geom_point(alpha = 0.6, color = "steelblue") +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
    geom_smooth(method = "loess", se = FALSE, color = "darkred") +
    labs(title = "Residuals vs Fitted: Model 1 (FEI + Inactivity)",
         x = "Fitted Values", y = "Residuals") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  p2 <- ggplot(model_data, aes(x = fitted2, y = residuals2)) +
    geom_point(alpha = 0.6, color = "darkgreen") +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
    geom_smooth(method = "loess", se = FALSE, color = "darkred") +
    labs(title = "Residuals vs Fitted: Model 2 (Inactivity + Food Access)",
         x = "Fitted Values", y = "Residuals") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  return(list(
    model1 = model1,
    model2 = model2,
    plot1 = p1,
    plot2 = p2,
    data = model_data
  ))
}
```

# ===== ADDITIONAL HELPER FUNCTIONS =====
```{r}
# Function to create a nice summary table for presentation
create_model_comparison_table <- function(model1, model2) {
  # Extract coefficients and statistics
  model1_coef <- summary(model1)$coefficients
  model2_coef <- summary(model2)$coefficients
  
  # Check if p-values are available (some robust regression summaries don't include them)
  has_pvalues <- ncol(model1_coef) >= 4 && ncol(model2_coef) >= 4
  
  if (has_pvalues) {
    # Create comparison table with p-values
    comparison <- data.frame(
      Model = c("Model 1: FEI + Inactivity", "", "", 
                "Model 2: Inactivity + Food Access", "", ""),
      Variable = c("Intercept", "Food Environment Index", "Physical Inactivity (%)",
                   "Intercept", "Physical Inactivity (%)", "Limited Food Access (%)"),
      Coefficient = c(round(model1_coef[,1], 3), round(model2_coef[,1], 3)),
      SE = c(round(model1_coef[,2], 3), round(model2_coef[,2], 3)),
      t_value = c(round(model1_coef[,3], 3), round(model2_coef[,3], 3)),
      p_value = c(ifelse(model1_coef[,4] < 0.001, "<0.001", round(model1_coef[,4], 3)),
                  ifelse(model2_coef[,4] < 0.001, "<0.001", round(model2_coef[,4], 3)))
    )
  } else {
    # Create comparison table without p-values
    comparison <- data.frame(
      Model = c("Model 1: FEI + Inactivity", "", "", 
                "Model 2: Inactivity + Food Access", "", ""),
      Variable = c("Intercept", "Food Environment Index", "Physical Inactivity (%)",
                   "Intercept", "Physical Inactivity (%)", "Limited Food Access (%)"),
      Coefficient = c(round(model1_coef[,1], 3), round(model2_coef[,1], 3)),
      SE = c(round(model1_coef[,2], 3), round(model2_coef[,2], 3)),
      t_value = c(round(model1_coef[,3], 3), round(model2_coef[,3], 3))
    )
  }
  
  return(comparison)
}

# Function to calculate model fit statistics
calculate_model_stats <- function(model1, model2, data) {
  # Calculate R-squared equivalent for robust regression
  model_data <- data %>%
    dplyr::select(percent_adults_with_obesity, 
           percent_physically_inactive,
           food_environment_index,
           percent_limited_access_to_healthy_foods) %>%
    na.omit()
  
  # Pseudo R-squared using correlation between fitted and actual values
  fitted1 <- fitted(model1)
  fitted2 <- fitted(model2)
  actual <- model_data$percent_adults_with_obesity
  
  r2_model1 <- cor(fitted1, actual)^2
  r2_model2 <- cor(fitted2, actual)^2
  
  # RMSE
  rmse1 <- sqrt(mean(residuals(model1)^2))
  rmse2 <- sqrt(mean(residuals(model2)^2))
  
  # MAE
  mae1 <- mean(abs(residuals(model1)))
  mae2 <- mean(abs(residuals(model2)))
  
  stats <- data.frame(
    Model = c("Model 1: FEI + Inactivity", "Model 2: Inactivity + Food Access"),
    Pseudo_R2 = round(c(r2_model1, r2_model2), 3),
    RMSE = round(c(rmse1, rmse2), 3),
    MAE = round(c(mae1, mae2), 3)
  )
  
  return(stats)
}

# Function to create regional comparison visualization
create_regional_comparison_viz <- function(data) {
  # Add region classification
  data_regional <- data %>%
    mutate(region = classify_region(state)) %>%
    filter(region != "Other") %>%
    dplyrselect(region, percent_adults_with_obesity, percent_physically_inactive, 
           food_environment_index, percent_limited_access_to_healthy_foods) %>%
    na.omit()
  
  # Create multiple comparison plots
  p1 <- ggplot(data_regional, aes(x = region, y = percent_adults_with_obesity, fill = region)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.2, fill = "white", alpha = 0.8) +
    labs(title = "Obesity Rates by Region", x = "Region", y = "Obesity Rate (%)") +
    theme_minimal() +
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
  
  p2 <- ggplot(data_regional, aes(x = region, y = percent_physically_inactive, fill = region)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.2, fill = "white", alpha = 0.8) +
    labs(title = "Physical Inactivity by Region", x = "Region", y = "Physical Inactivity (%)") +
    theme_minimal() +
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
  
  p3 <- ggplot(data_regional, aes(x = region, y = food_environment_index, fill = region)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.2, fill = "white", alpha = 0.8) +
    labs(title = "Food Environment Index by Region", x = "Region", y = "FEI Score") +
    theme_minimal() +
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
  
  return(list(obesity = p1, inactivity = p2, fei = p3))
}

# Function to identify extreme counties for discussion
identify_extreme_counties <- function(data) {
  extreme_counties <- data %>%
    select(state, county, percent_adults_with_obesity, percent_physically_inactive, 
           food_environment_index, percent_limited_access_to_healthy_foods) %>%
    na.omit() %>%
    mutate(
      obesity_percentile = ntile(percent_adults_with_obesity, 100),
      inactivity_percentile = ntile(percent_physically_inactive, 100),
      fei_percentile = ntile(food_environment_index, 100)
    ) %>%
    filter(
      obesity_percentile >= 95 | obesity_percentile <= 5 |
      inactivity_percentile >= 95 | inactivity_percentile <= 5 |
      fei_percentile >= 95 | fei_percentile <= 5
    ) %>%
    arrange(desc(obesity_percentile))
  
  return(extreme_counties)
}
```

# ===== EXAMPLE USAGE =====
```{r}
# Uncomment and run these functions with your data

# Slide 6: Obesity Map
obesity_map <- create_obesity_map(data_2025)
print(obesity_map)

# Slide 7: Scatterplot Analysis
inactivity_plot <- create_inactivity_plot(data_2025)
food_access_plot <- create_food_access_plot(data_2025)
print(inactivity_plot)
print(food_access_plot)

# Slide 8: Food Environment Index
fei_plot <- create_fei_plot(data_2025)
print(fei_plot)

# Slide 9: Variable Importance
fei_importance <- create_fei_importance(data_2025)
print(fei_importance)

# Slide 10: Correlation Matrix
create_correlation_matrix(data_2025)

# Slide 11: County-Level Analysis
county_stats <- create_county_summary(data_2025)
create_scatter_matrix(data_2025)
extreme_counties <- identify_extreme_counties(data_2025)

# Slide 12: Regional Analysis
regional_analysis <- create_regional_analysis(data_2025)
regional_viz <- create_regional_comparison_viz(data_2025)
print(regional_analysis$summary)
print(regional_viz$obesity)
print(regional_viz$inactivity)
print(regional_viz$fei)

# ===== SAFE EXECUTION WRAPPER =====
# Use this function to safely run your analysis
run_slide_analysis <- function(data_2025, slide_number) {
  
  switch(as.character(slide_number),
    "6" = {
      cat("Creating Obesity Map...\n")
      tryCatch({
        obesity_map <- create_obesity_map(data_2025)
        print(obesity_map)
      }, error = function(e) cat("Error in slide 6:", e$message, "\n"))
    },
    
    "7" = {
      cat("Creating Scatterplot Analysis...\n")
      tryCatch({
        inactivity_plot <- create_inactivity_plot(data_2025)
        food_access_plot <- create_food_access_plot(data_2025)
        print(inactivity_plot)
        print(food_access_plot)
      }, error = function(e) cat("Error in slide 7:", e$message, "\n"))
    },
    
    "8" = {
      cat("Creating Food Environment Index Plot...\n")
      tryCatch({
        fei_plot <- create_fei_plot(data_2025)
        print(fei_plot)
      }, error = function(e) cat("Error in slide 8:", e$message, "\n"))
    },
    
    "9" = {
      cat("Creating Variable Importance Analysis...\n")
      tryCatch({
        fei_importance <- create_fei_importance(data_2025)
        print(fei_importance)
      }, error = function(e) cat("Error in slide 9:", e$message, "\n"))
    },
    
    "10" = {
      cat("Creating Correlation Matrix...\n")
      tryCatch({
        create_correlation_matrix(data_2025)
      }, error = function(e) cat("Error in slide 10:", e$message, "\n"))
    },
    
    "11" = {
      cat("Creating County-Level Analysis...\n")
      tryCatch({
        county_stats <- create_county_summary(data_2025)
        print(county_stats)
        create_scatter_matrix(data_2025)
        extreme_counties <- identify_extreme_counties(data_2025)
        print(head(extreme_counties, 10))
      }, error = function(e) cat("Error in slide 11:", e$message, "\n"))
    },
    
    "12" = {
      cat("Creating Regional Analysis...\n")
      tryCatch({
        regional_analysis <- create_regional_analysis(data_2025)
        regional_viz <- create_regional_comparison_viz(data_2025)
        print(regional_analysis$summary)
        print(regional_viz$obesity)
        print(regional_viz$inactivity)
        print(regional_viz$fei)
      }, error = function(e) cat("Error in slide 12:", e$message, "\n"))
    },
    
    "13" = {
      cat("Creating Huber Regression Models...\n")
      tryCatch({
        huber_results <- create_huber_models(data_2025)
        
        # Print model summaries
        cat("\n=== MODEL 1 SUMMARY ===\n")
        print(summary(huber_results$model1))
        
        cat("\n=== MODEL 2 SUMMARY ===\n")
        print(summary(huber_results$model2))
        
        # Create comparison table
        model_comparison <- create_model_comparison_table(huber_results$model1, huber_results$model2)
        cat("\n=== MODEL COMPARISON TABLE ===\n")
        print(model_comparison)
        
        # Calculate model statistics
        model_stats <- calculate_model_stats(huber_results$model1, huber_results$model2, data_2025)
        cat("\n=== MODEL FIT STATISTICS ===\n")
        print(model_stats)
        
        # Print residual plots
        print(huber_results$plot1)
        print(huber_results$plot2)
        
      }, error = function(e) cat("Error in slide 13:", e$message, "\n"))
    },
    
    cat("Invalid slide number. Use 6-13\n")
  )
}
```
# ===== EXAMPLE USAGE - SAFE EXECUTION =====
```{r}
# Run individual slides safely:
 run_slide_analysis(data_2025, "6")   # Obesity map
 run_slide_analysis(data_2025, "7")   # Scatterplots
 run_slide_analysis(data_2025, "8")   # FEI analysis
 run_slide_analysis(data_2025, "9")   # Variable importance
 run_slide_analysis(data_2025, "10")  # Correlation matrix
 run_slide_analysis(data_2025, "11")  # County analysis
 run_slide_analysis(data_2025, "12")  # Regional comparison
 run_slide_analysis(data_2025, "13")  # Huber regression

# Or run all slides:
 for(i in 6:13) {
   cat("\n", rep("=", 50), "\n")
   cat("SLIDE", i, "\n")
   cat(rep("=", 50), "\n")
   run_slide_analysis(data_2025, as.character(i))
 }
```