---
title: "Mapping Obesity: A County-Level Analysis of Lifestyle and Food Environment Factors"
subtitle: "SURE 2025"
author:
  - "<strong>Helen Peng</strong><br><small>Carnegie Mellon University</small>"
  - "<strong>Jordan Faustin</strong><br><small>Howard University</small>"
  - "<strong>Anvith Thumma</strong><br><small>Arizona State University</small>"
format: 
  revealjs: 
    embed-resources: true
    theme: serif
    slide-number: true
    smaller: true
    include-after-body: |
      <style>
        .reveal .slides section.title-slide {
          text-align: center;
        }
        .reveal .slides section.title-slide .author p {
          margin: 1em 0;
          line-height: 1.5;
        }
      </style>
editor: visual
---

## Significance

-   **Public Health Implications:** <br> Obesity is a major contributor to rising healthcare costs and chronic diseases such as diabetes, heart disease, and cancer.

-   **Social Factors of Health:** <br> Physical Inactivity and food access are modifiable factors linked to obesity. In turn, studying their impact can highlight how our environment and infrastructure, like our parks, grocery stores, and schools, can shape our health outcomes.

-   **Policy and Community Planning:** <br> New information revealed through findings can help to make informed policies for urban planning in under-served areas and school/workplace wellness programs.

## Does physical inactivity level and access to healthy food affect obesity rate at the county level? {.center}

## Agenda

-   **Data Overview**
-   **Initial Analysis**
-   **Understanding Direction**
-   **Modeling and Comparison**
-   **Discussion**
-   **Recommendations**
-   **Appendix**

## Data Overview

-   **Observations:**

    -   Data at the county level across all 50 U.S. states in 2025

-   **Variables of Interest:**

    -   \% Adults with Obesity

    -   \% Physically Inactive

    -   Food Environment Index (1-10)

    -   State and County Identifiers

-   **Handling Missing Data:**

    -   If data on Food Environment Index or Physical Inactivity was missing for a county we used the state average.

    -   Seven counties that did not have data for obesity rate, our response variable and so, we dropped those counties to do a complete case analysis.

```{r setup}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center",
               echo=FALSE, tidy=TRUE)
```

```{r cleaning}
library(readxl)
library(tidyverse)
theme_set(theme_bw())

data_2025 <- read_excel("C:/Users/jorda/Desktop/Capstone Obesity/2025_CHRD.xlsx", sheet = "Additional Measure Data", skip = 1)

food_index_2025 <- read_excel("C:/Users/jorda/Desktop/Capstone Obesity/2025_CHRD.xlsx", sheet = "Select Measure Data", skip = 1)

data_2025 <- data_2025 |> 
  janitor::clean_names() 

food_index_2025 <- food_index_2025 |> 
  janitor::clean_names() 

data_2025 <- merge(data_2025, food_index_2025, by = "fips") 
data_2025 <- data_2025[, !grepl("percent_ci", names(data_2025))]
data_2025 <- data_2025[, !grepl("national_z", names(data_2025))]
data_2025 <- data_2025[, !grepl("injury_death", names(data_2025))]
data_2025 <- data_2025[, !grepl("percentile", names(data_2025))]
data_2025 <- data_2025[, !grepl("_white", names(data_2025))]
data_2025 <- data_2025[, !grepl("_hispanic", names(data_2025))]
data_2025 <- data_2025[, !grepl("_native", names(data_2025))]
data_2025 <- data_2025[, !grepl("_aian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_asian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_black", names(data_2025))]
data_2025 <- data_2025[, !grepl("dentist", names(data_2025))]
data_2025 <- data_2025[, !grepl("unreliable", names(data_2025))]
data_2025 <- data_2025[, !grepl("deaths", names(data_2025))]

drop_cols <- c("child_mortality_rate","hiv_prevalence_rate", " suicide_rate_age_adjusted", "crude_rate","teen_birth_rate","chlamydia_rate","number_chlamydia_cases", "drug_overdose_mortality_rate","other_primary_care_provider_rate", "other_primary_care_provider_ratio","percent_census_participation", "average_grade_performance_252", " average_grade_performance_246","cohort_size","gender_pay_gap","segregation_index_287","homicide_rate","motor_vehicle_mortality_rate", "percent_disconnected_youth","firearm_fatalities_rate","percent_lacking_support", "percent_low_birth_weight","number_primary_care_physicians","primary_care_physicians_ratio","primary_care_physicians_rate","number_mental_health_providers", "mental_health_provider_rate", "mental_health_provider_ratio", "visits_per_service_area_population", "presence_of_water_violation","average_daily_pm2_5", "state.y","county.y", "percent_with_annual_mammogram", "preventable_hospitalization_rate", "years_of_potential_life_lost_rate","number_firearm_fatalities","child_care_centers_per_1_000_children", "number_firearm_fatalities", "number_child_care_centers","spending_per_pupil","segregation_index_258","high_school_graduation_rate","average_grade_performance_246","disaster_declarations","weeks_in_moderate_or_greater_drought", "life_expectancy", "severe_housing_cost_burden", "percent_adults_with_diabetes","suicide_rate_age_adjusted","percent_fair_or_poor_health", "percent_voter_turnout", "percent_with_disability", "age_adjusted_death_rate", "percent_drive_alone_to_work", "income_ratio", "mens_median_earnings", "womens_median_earnings", "infant_mortality_rate","percent_uninsured_children","percent_uninsured_adults","percent_with_access_to_exercise_opportunities")

data_2025 <- data_2025[,!(names(data_2025) %in% drop_cols)]

data_2025 <- data_2025  |> 
  rename(
    county = county.x,
    state = state.x
  )

old_ct_fips <- c("09001", "09003", "09005", "09007", "09009", "09011", "09013", "09015")  

statewide_food_index <- data_2025 |> 
  filter(is.na(county) & !is.na(food_environment_index)) |> 
  dplyr::select(state, food_environment_index) |> 
  rename(statewide_food_environment_index = food_environment_index)

data_2025 <- data_2025 |> 
  filter(!(fips %in% old_ct_fips)) |> 
  left_join(statewide_food_index, by = "state") |> 
  mutate(
    food_environment_index = if_else(
      is.na(food_environment_index) & !is.na(county),
      statewide_food_environment_index,
      food_environment_index),
      days_above_90f = if_else(days_above_90f > 365, 365, days_above_90f)) |> 
  dplyr::select(-statewide_food_environment_index) |> 
  filter(!is.na(percent_adults_with_obesity))
```

## As Physical Inactivity increases Obesity Rates increase as well, and as the Food Environment Index increase Obesity Rates start to decline {style="font-size: 0.4em;"}

```{r scatterplot}
library(patchwork)

physical_obesity_plot <- data_2025 |>
  ggplot(aes(x = percent_physically_inactive, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "% Physically Inactive",
    y = "% Adults with Obesity"
  ) 

food_obesity_plot <- data_2025 |>
  ggplot(aes(x = food_environment_index, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "Food Environment Index (0 = worst, 10 = best)",
    y = "% Adults with Obesity"
  ) 
combined_plot <- physical_obesity_plot + food_obesity_plot

# Save with larger dimensions
ggsave("combined_plot.png", combined_plot, width = 20, height = 15, dpi = 800)

combined_plot
```

## Obesity Rates are notably higher in the Midwest and South compared to the Northeast and West {style="font-size: 0.4em;"}

```{r obesity_map}
library(usmap)
library(viridis)
library(ggplot2)

# Define fixed limits based on your data range (18 to 53)
legend_min <- 18
legend_max <- 53

# Create quantile breaks (optional, but helps distribute colors)
q_brks <- floor(quantile(
  data_2025$percent_adults_with_obesity,
  probs = seq(0, 1, length.out = 5),
  na.rm = TRUE
))

# Plot map with FIXED legend scaling
obesity_map <- plot_usmap(data = data_2025, values = "percent_adults_with_obesity", color = NA) +
  scale_fill_viridis_c(
    option = "inferno",     # Blue-yellow color scale (colorblind-friendly)
    direction = -1,         # Reverse to emphasize high values
    breaks = q_brks,        # Quantile breaks
    labels = round(q_brks, 1),
    name = "Obesity Rate (%)",
    na.value = "grey90",
    limits = c(legend_min, legend_max)  # FIXED SCALE (18-53)
  ) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "bold")
  )

# Save with larger dimensions (width & height in inches)
ggsave("obesity_map.png", obesity_map, width = 20, height = 15, dpi = 800)

obesity_map
```

## Fuel or Fire {.center}

There is a relationship between obesity rates, physical inactivity, and healthy food environments. Now we want to confirm whether we're looking at the "Fuel" or the "Fire" of the situation. Reasoning behind this process is to clarify that we are targeting a driver and not a symptom.

## Surprisingly, the Food Environment Index does not rank among the top predictors

```{r vip}

# CART-algorthim regression tree 10 fold CV
library(caret)
library(vip)

set.seed(10)
data_2025 <- data_2025 |> na.omit()

tree <- train(percent_adults_with_obesity ~ . - fips - state - county,
              method = "rpart", tuneLength = 20,
              trControl = trainControl(method = "cv", number = 10),
              data = data_2025)

top_vi <- head(vi(tree), 15)

pretty_labels <- c(
  "percent_physically_inactive" = "Physically Inactive",
  "median_household_income" = "Median Household Income",
  "percent_adults_reporting_currently_smoking" = "Smoking",
  "percent_excessive_drinking" = "Excessive Drinking",
  "percent_below_18_years_of_age" = "Population Below 18 Years Old",
  "percent_insufficient_sleep" = "Insufficient Sleep",
  "school_funding_adequacy" = "School Funding Adequacy",
  "percent_households_with_severe_cost_burden" = "Severe Cost Burden",
  "percent_uninsured_adults" = "Uninsured Adults",
  "traffic_volume" = "Traffic Volume",
  "food_environment_index" = "Food Environment Index",
  "percent_uninsured" = "Uninsured Adults"
)

top_vi$Variable <- ifelse(
  top_vi$Variable %in% names(pretty_labels),
  pretty_labels[top_vi$Variable],
  top_vi$Variable
)

vip(top_vi, aesthetics = list(fill = "steelblue")) +
  labs(y = "Variable Importance (Total Reduction in RSS)")

```

## Modeling and Comparison

```{r bivariate_model}
# Create model comparison summary table
library(MASS)
library(knitr)

# Prepare data and fit models
model_data_2025 <- data_2025 |> 
  mutate(
    obesity_prop = percent_adults_with_obesity / 100,
    logit_obesity = log(obesity_prop / (1 - obesity_prop))
  )

lm_simple <- lm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
    data = model_data_2025)

huber_model <- rlm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
    data = model_data_2025)

# Create model comparison table
model_comparison <- data.frame(
  Model = c("Linear Model (OLS)", "Huber Robust Model"),
  `Food Environment Index` = c(
    sprintf("%.4f (%.4f)", 
            coef(lm_simple)[2], 
            summary(lm_simple)$coefficients[2, 2]),
    sprintf("%.4f (%.4f)", 
            coef(huber_model)[2], 
            summary(huber_model)$coefficients[2, 2])
  ),
  `Physically Inactive` = c(
    sprintf("%.4f (%.4f)", 
            coef(lm_simple)[3], 
            summary(lm_simple)$coefficients[3, 2]),
    sprintf("%.4f (%.4f)", 
            coef(huber_model)[3], 
            summary(huber_model)$coefficients[3, 2])
  ),
  `R-squared` = c(
    sprintf("%.4f", summary(lm_simple)$r.squared),
    "N/A"
  ),
  `Residual SE` = c(
    sprintf("%.4f", summary(lm_simple)$sigma),
    sprintf("%.4f", huber_model$s)
  )
)

# Display table
model_comparison |> 
  kable(
    digits = 4,
    col.names = c("Model", "Food Environment Index (SE)", "Physically Inactive (SE)", 
                  "R-squared", "Residual SE"),
    caption = "Comparison of Linear and Huber Robust Models for Obesity Prediction"
  )
```

## Key Takeaways

-   **Physical Inactivity** is the strongest and most consistent predictor of obesity.

-   **Food Insecurity Percentage** becomes a statistically significant predictor once we control for physical inactivity and drop FEI.

-   **Access to Healthy Food Percentage** is not statistically significant in any specification we tested.

-   Relying on the FEI alone is misleading because its signal disappears once key components (and physical inactivity) are included.

## Recommendations

#### Prioritize Physical Activity Infrastructure

-   Building walkable Communities (expanding sidewalks, bike lanes, and trails in high-obesity counties)
-   Funding Safe Routes to School programs to improve sidewalks, crosswalks, and safety patrols.

#### Improve Food Access with Targeted Interventions

-   Deploying Mobile Food Solutions such as produce vans (e.g., Veggie Van in NC) in food deserts. and hosting pop-up markets paired with nutrition workshops.

#### Redirecting Funding

-   Shifting resources from FEI-based programs to hyper-local initiatives (e.g., subsidized CSA boxes for low-income families).

## Possible Future Work

-   Establishing a count of parks and wellness centers per county

-   Mapping their respective distance and location

-   Checking for trends on when individuals are going to the fitness locations

## Questions? {.center, style="font-size: 1.5em;"}

## Thank You {.center, style="font-size: 1.5em;"}

## Appendix

-   Correlation Matrix
-   Diagnostic Plots for Linear Regression Models
-  
##Apendix
### Food Environment Index appears to be a confounding variable {.smaller}

```{r}
# Clean data
set.seed(10)
data_2025 <- data_2025 |> na.omit()

tree_fei <- train(food_environment_index ~ . - fips - state - county,
              method = "rpart", tuneLength = 20,
              trControl = trainControl(method = "cv", number = 10),
              data = data_2025)

# Top 10 predictors of FEI
top_vi_fei <- head(vi(tree_fei), 10)

pretty_labels_fei <- c(
  "percent_limited_access_to_healthy_foods" = "Limited Access to Healthy Foods (%)",
  "percent_food_insecure" = "Food Insecurity (%)",
  "median_household_income" = "Median Household Income",
  "percent_children_in_poverty" = "Children in Poverty (%)",
  "number_limited_access_to_healthy_foods" = "Number with Limited Access to Healthy Foods",
  "average_number_of_physically_unhealthy_days" = "Avg. Physically Unhealthy Days",
  "percent_frequent_physical_distress" = "Frequent Physical Distress (%)",
  "percent_physically_inactive" = "Physically Inactive (%)",
  "number_completed_high_school" = "Completed High School",
  "population_173" = "Population "
)

top_vi_fei$Variable <- ifelse(
  top_vi_fei$Variable %in% names(pretty_labels_fei),
  pretty_labels_fei[top_vi_fei$Variable],
  top_vi_fei$Variable
)

vip(top_vi_fei, aesthetics = list(fill = "darkgreen")) +
  labs(y = "Variable Importance (Total Reduction in RSS)")

```

```{r, eval=FALSE}
summary(lm(logit_obesity ~ food_environment_index, data = data_2025))

summary(lm(logit_obesity ~ food_environment_index + percent_limited_access_to_healthy_foods + percent_food_insecure, data = data_2025))

summary(lm(logit_obesity ~ percent_physically_inactive + percent_limited_access_to_healthy_foods + percent_food_insecure, data = data_2025))

```


## Appendix

```{r linear model}
# Transform obesity to logit scale (as shown in slides)
data_2025 <- data_2025 |> 
  mutate(
    obesity_prop = percent_adults_with_obesity / 100,
    logit_obesity = log(obesity_prop / (1 - obesity_prop))
  )

lm_base <- lm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
              data = data_2025)

library(MASS)
huber_model <- rlm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
                   data = data_2025)

lm_expanded <- lm(logit_obesity ~ percent_physically_inactive + 
                  percent_limited_access_to_healthy_foods + 
                  percent_food_insecure,
                data = data_2025)

# Compare models
library(broom)
library(knitr)

model_comparison <- list(
  "Base Model" = tidy(lm_base),
  "Huber Robust" = tidy(huber_model),
  "Expanded Model" = tidy(lm_expanded)
)

kable(model_comparison, digits = 4, 
      caption = "Comparison of Obesity Prediction Models")

# Check assumptions
par(mfrow = c(2, 2))
plot(lm_base)

# Check for multicollinearity
car::vif(lm_expanded)

# Check residual distribution
ggplot(data.frame(resid = residuals(lm_base)), aes(sample = resid)) +
  stat_qq() + stat_qq_line()
```

<!-- 1. Residuals vs. Fitted Values Plot -->

<!-- X-axis: Fitted Values (Predicted Obesity Rates) -->

<!-- Y-axis: Residuals (Actual - Predicted Obesity Rates) -->

<!-- What it checks: -->

<!-- Whether your model misses patterns (non-linearity) -->

<!-- If residuals have constant variance (homoscedasticity) -->

<!-- Good: -->

<!-- Points randomly scattered around zero (like a cloud) -->

<!-- No obvious curves or funnels -->

<!-- Bad: -->

<!-- U-shaped curve → Missing a non-linear term (e.g., maybe inactivity only affects obesity above a certain threshold) -->

<!-- Funnel shape → Variance changes with predictions (e.g., obesity harder to predict in very urban counties) -->

<!-- For obesity study: -->

<!-- If high-obesity counties (right side) have bigger residuals, maybe the model works better for average counties than extreme ones. -->

<!-- 2. Q-Q Plot (Theoretical Quantiles vs. Standardized Residuals) -->

<!-- X-axis: Theoretical Normal Quantiles (Expected pattern if normal) -->

<!-- Y-axis: Standardized Residuals (Actual errors, scaled) -->

<!-- What it checks: -->

<!-- Whether residuals follow a normal distribution -->

<!-- Good: -->

<!-- Points follow the straight diagonal line -->

<!-- Bad: -->

<!-- S-curve → Heavy tails (extreme outliers) -->

<!-- Flat middle → Too many average errors -->

<!-- Points veer off at ends → Extreme outliers -->

<!-- For obesity study: -->

<!-- If tails bend up, your model underestimates obesity in the worst counties. If bends down, it overestimates. -->

<!-- 3. Scale-Location Plot (Fitted Values vs. √Standardized Residuals) -->

<!-- X-axis: Fitted Values (Predicted Obesity Rates) -->

<!-- Y-axis: √|Standardized Residuals| (Square root of absolute errors) -->

<!-- What it checks: -->

<!-- Whether residual variance is consistent (homoscedasticity) -->

<!-- Good: -->

<!-- Flat red line (no trend in variance) -->

<!-- Bad: -->

<!-- Upward slope → More error as predictions increase (e.g., obesity harder to predict in high-obesity counties) -->

<!-- Downward slope → Opposite issue -->

<!-- For obesity study: -->

<!-- If high-fitted values show more spread, consider transforming obesity rates or adding predictors for extreme counties. -->

<!-- 4. Residuals vs. Leverage Plot (Leverage vs. Standardized Residuals) -->

<!-- X-axis: Leverage (How unusual a county's predictors are) -->

<!-- Y-axis: Standardized Residuals (How badly the model predicted it) -->

<!-- What it checks: -->

<!-- Influential outliers that distort the model -->

<!-- Key terms: -->

<!-- Leverage: Measures how extreme a county's predictors are (e.g., super high inactivity + low fast food) -->

<!-- Cook’s Distance (red curves): Shows how much removing a county changes the model -->

<!-- Danger zones: -->

<!-- Top-right: High leverage + high residual (badly predicted unusual cases) -->

<!-- Bottom-right: High leverage but well-predicted (pulls the model toward it) -->

<!-- For obesity study: -->

<!-- A wealthy county with bad food access but low obesity might sit here. -->

<!-- Rural counties with unique trends could dominate the model unfairly. -->
