---
title: "Mapping Obesity: A County-Level Analysis of Lifestyle and Food Environment Factors"
subtitle: "SURE 2025"
author:
  - "<strong>Helen Peng</strong><br><small>Carnegie Mellon University</small>"
  - "<strong>Jordan Faustin</strong><br><small>Howard University</small>"
  - "<strong>Anvith Thumma</strong><br><small>Arizona State University</small>"
format: 
  revealjs:
    theme: beige
    slide-number: true
    smaller: true
    include-after-body: |
      <style>
        .reveal .slides section.title-slide {
          text-align: center;
        }
        .reveal .slides section.title-slide .author p {
          margin: 1em 0;
          line-height: 1.5;
        }
      </style>
editor: visual
---


## Does physical inactivity level and access to healthy food affect obesity rate at the county level? {.center}


## Why This Question is Important

-   **Public Health Implications:** 
Obesity is a major contributor to rising healthcare costs and chronic diseases such as diabetes, heart disease, and cancer.

-   **Social Factors of Health:** 
Physical Inactivity and food access are modifiable factors linked to obesity. In turn, studying their impact highlights how our environment and infrastructure, like parks, grocery stores, and schools, can shape our health outcomes.

-   **Policy and Community Planning:** 
New information made through findings can inform policies like urban planning, subsidizing healthy food retailers in under-served areas, school/workplace wellness programs.


## Agenda

-   **Data Overview**
-   **Initial Analysis**
-   **Percentage Distribution**
-   **Predictor Analysis**
-   **Regression Model**
-   **Model Comparison**
-   **Recommendations**
-   **Appendix**


## Data Overview

-   **Observations:** Data at the county level across all 50 U.S. states in 2025

-   **Variables of Interest:**

    -   \% Adults with Obesity

    -   \% Physically Inactive

    -   Food Environment Index (1-10) - Higher is Better

    -   State and County Identifiers

    -   Days above 90°F
    
    -   Access to Parks

-   **Handling Missing Data:**

    -   If data on Food Environment Index or Physical Inactivity was missing for a county we used the state average.

    -   Seven counties that did not have data for obesity rate, our response variable and so, we dropped those counties to do a complete case analysis.

```{r setup}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center",
               echo=FALSE, tidy=TRUE)
```

```{r cleaning}
library(readxl)
library(tidyverse)
theme_set(theme_bw())

data_2025 <- read_excel("C:/Users/jorda/Desktop/Capstone Obesity/2025_CHRD.xlsx", sheet = "Additional Measure Data", skip = 1)

food_index_2025 <- read_excel("C:/Users/jorda/Desktop/Capstone Obesity/2025_CHRD.xlsx", sheet = "Select Measure Data", skip = 1)

data_2025 <- data_2025 |> 
  janitor::clean_names() 

food_index_2025 <- food_index_2025 |> 
  janitor::clean_names() 

data_2025 <- merge(data_2025, food_index_2025, by = "fips") 
data_2025 <- data_2025[, !grepl("percent_ci", names(data_2025))]
data_2025 <- data_2025[, !grepl("national_z", names(data_2025))]
data_2025 <- data_2025[, !grepl("injury_death", names(data_2025))]
data_2025 <- data_2025[, !grepl("percentile", names(data_2025))]
data_2025 <- data_2025[, !grepl("_white", names(data_2025))]
data_2025 <- data_2025[, !grepl("_hispanic", names(data_2025))]
data_2025 <- data_2025[, !grepl("_native", names(data_2025))]
data_2025 <- data_2025[, !grepl("_aian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_asian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_black", names(data_2025))]
data_2025 <- data_2025[, !grepl("dentist", names(data_2025))]
data_2025 <- data_2025[, !grepl("unreliable", names(data_2025))]
data_2025 <- data_2025[, !grepl("deaths", names(data_2025))]

drop_cols <- c("child_mortality_rate","hiv_prevalence_rate", " suicide_rate_age_adjusted", "crude_rate","teen_birth_rate","chlamydia_rate","number_chlamydia_cases", "drug_overdose_mortality_rate","other_primary_care_provider_rate", "other_primary_care_provider_ratio","percent_census_participation", "average_grade_performance_252", " average_grade_performance_246","cohort_size","gender_pay_gap","segregation_index_287","homicide_rate","motor_vehicle_mortality_rate", "percent_disconnected_youth","firearm_fatalities_rate","percent_lacking_support", "percent_low_birth_weight","number_primary_care_physicians","primary_care_physicians_ratio","primary_care_physicians_rate","number_mental_health_providers", "mental_health_provider_rate", "mental_health_provider_ratio", "visits_per_service_area_population", "presence_of_water_violation","average_daily_pm2_5", "state.y","county.y", "percent_with_annual_mammogram", "preventable_hospitalization_rate", "years_of_potential_life_lost_rate","number_firearm_fatalities","child_care_centers_per_1_000_children", "number_firearm_fatalities", "number_child_care_centers","spending_per_pupil","segregation_index_258","high_school_graduation_rate","average_grade_performance_246","disaster_declarations","weeks_in_moderate_or_greater_drought", "life_expectancy", "severe_housing_cost_burden", "percent_adults_with_diabetes","suicide_rate_age_adjusted","percent_fair_or_poor_health", "percent_voter_turnout", "percent_with_disability", "age_adjusted_death_rate", "percent_drive_alone_to_work", "income_ratio", "mens_median_earnings", "womens_median_earnings", "infant_mortality_rate","percent_uninsured_children","percent_uninsured_adults","percent_with_access_to_exercise_opportunities")

data_2025 <- data_2025[,!(names(data_2025) %in% drop_cols)]

data_2025 <- data_2025  |> 
  rename(
    county = county.x,
    state = state.x
  )

old_ct_fips <- c("09001", "09003", "09005", "09007", "09009", "09011", "09013", "09015")  

statewide_food_index <- data_2025 |> 
  filter(is.na(county) & !is.na(food_environment_index)) |> 
  dplyr::select(state, food_environment_index) |> 
  rename(statewide_food_environment_index = food_environment_index)

data_2025 <- data_2025 |> 
  filter(!(fips %in% old_ct_fips)) |> 
  left_join(statewide_food_index, by = "state") |> 
  mutate(
    food_environment_index = if_else(
      is.na(food_environment_index) & !is.na(county),
      statewide_food_environment_index,
      food_environment_index),
      days_above_90f = if_else(days_above_90f > 365, 365, days_above_90f)) |> 
  dplyr::select(-statewide_food_environment_index) |> 
  filter(!is.na(percent_adults_with_obesity))
```

## There is a positive correlation between physical inactivity and obesity rates, and a negative correlation between healthy food access

```{r scatterplot}
library(patchwork)

physical_obesity_plot <- data_2025 |>
  ggplot(aes(x = percent_physically_inactive, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "% Physically Inactive",
    y = "% Adults with Obesity"
  ) +
  theme_minimal()

food_obesity_plot <- data_2025 |>
  ggplot(aes(x = food_environment_index, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "Food Environment Index (0 = worst, 10 = best)",
    y = "% Adults with Obesity"
  ) +
  theme_minimal()

physical_obesity_plot + food_obesity_plot 
```

## CDC data confirms obesity rates are significantly higher in the Midwest and South compared to the Northeast and West

```{r obesity_map}
library(usmap)
library(viridis)
library(ggplot2)

# Define fixed limits based on your data range (18 to 53)
legend_min <- 18
legend_max <- 53

# Create quantile breaks (optional, but helps distribute colors)
q_brks <- floor(quantile(
  data_2025$percent_adults_with_obesity,
  probs = seq(0, 1, length.out = 5),
  na.rm = TRUE
))

# Plot map with FIXED legend scaling
plot_usmap(data = data_2025, values = "percent_adults_with_obesity", color = NA) +
  scale_fill_viridis_c(
    option = "cividis",     # Blue-yellow color scale (colorblind-friendly)
    direction = -1,         # Reverse to emphasize high values
    breaks = q_brks,        # Quantile breaks
    labels = round(q_brks, 1),
    name = "Obesity Rate (%)",
    na.value = "grey90",
    limits = c(legend_min, legend_max)  # FIXED SCALE (18-53)
  ) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "bold")
  )
```

## Fuel or Fire (CART)
- **We understand that there are correlations between obesity rates, physical inactivity, and healthy food environments. Now we to confirm whether we're looking at the "Fuel" or the "Fire" of the situation**
- **Importance behind this process is that it assists us with targeting the root cause**

## Surprisingly, the Food Environment Index does not rank among the top predictors of county-level obesity rates.

```{r vip}

# CART-algorthim regression tree 10 fold CV
library(caret)
library(vip)

set.seed(10)
data_2025 <- data_2025 |> na.omit()

tree <- train(percent_adults_with_obesity ~ . - fips - state - county,
              method = "rpart", tuneLength = 20,
              trControl = trainControl(method = "cv", number = 10),
              data = data_2025)

top_vi <- head(vi(tree), 15)

pretty_labels <- c(
  "percent_physically_inactive" = "Physically Inactive",
  "median_household_income" = "Median Household Income",
  "percent_adults_reporting_currently_smoking" = "Smoking",
  "percent_excessive_drinking" = "Excessive Drinking",
  "percent_below_18_years_of_age" = "Population Below 18 Years Old",
  "percent_insufficient_sleep" = "Insufficient Sleep",
  "school_funding_adequacy" = "School Funding Adequacy",
  "percent_households_with_severe_cost_burden" = "Severe Cost Burden",
  "percent_uninsured_adults" = "Uninsured Adults",
  "traffic_volume" = "Traffic Volume",
  "food_environment_index" = "Food Environment Index",
  "percent_uninsured" = "Uninsured"
)

top_vi$Variable <- ifelse(
  top_vi$Variable %in% names(pretty_labels),
  pretty_labels[top_vi$Variable],
  top_vi$Variable
)

vip(top_vi, aesthetics = list(fill = "steelblue")) +
  labs(y = "Variable Importance (Total Reduction in RSS)")

```

## Linear Regression Model

```{r bivariate_model}
library(MASS)

data_2025 <- data_2025 |> 
  mutate(
    obesity_prop = percent_adults_with_obesity / 100,
    logit_obesity = log(obesity_prop / (1 - obesity_prop))
  )

lm_simple <- lm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
    data = data_2025)

summary(lm_simple)
# plot(lm_simple) -> Moved to appendix

huber_model <- rlm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
    data = data_2025)

summary(huber_model)
# Need to justify this over lm_simple
```

```{r multivariate_model}

```

## Model Comparsion

## Recommendations

-   Prioritize the young population by targeting environments that shape long-term health behaviors early in life.

-   Enhance sidewalk infrastructure across all clusters to promote walkability, encourage physical activity, and support safe routes for children and families.

-   Invest in school funding to strengthen the local food environment index and promote healthy habits—such as walking to school and accessing nutritious, affordable meals.

## Appendix

-   Correlation Matrix
-   Diagnostic Plots for Linear Regression Models
-   
