---
title: "Does physical inactivity level and access to healthy food affect obesity rate at the county level?"
author: Helen Peng, Anvith Thumma, and Jordan Faustin
format: 
  revealjs:
    theme: simple
    slide-number: true
    smaller: true
editor: visual
---

## Why This Question is Important

-   **Public Health Crisis:** Obesity is a major contributor to chronic diseases (diabetes, heart disease, cancer) and rising healthcare costs.

-   **Disparities in Obesity:** Obesity rates vary widely by region, income, and race/ethnicity. Identifying structural barriers (like food access and inactivity) can inform equitable policies.

-   **Rise in Weight Loss Drugs:** Medications like Ozempic and Wegovy are gaining popularity as quick fixes, but they’re expensive and don’t address root causes like environment or behavior.

## Data Overview

-   **Observations:** Data at the county level across all 50 U.S. states in 2025

-   **Variables of Interest:**

    -   \% Adults with Obesity

    -   \% Physically Inactive

    -   Food Environment Index (1-10) - Higher is Better

    -   State and County Identifiers

-   **Handling Missing Data:**

    -   If data on Food Environment Index or Physical Inactivity was missing for a county we used the state average.

    -   Seven counties that did not have data for obesity rate, our response variable and so, we dropped those counties to do a complete case analysis.

```{r setup}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center",
               echo=FALSE, tidy=TRUE)
```

```{r cleaning}
library(readxl)
library(tidyverse)
theme_set(theme_bw())

data_2025 <- read_excel("C:/Users/jorda/Desktop/Capstone Obesity/2025_CHRD.xlsx", sheet = "Additional Measure Data", skip = 1)

food_index_2025 <- read_excel("C:/Users/jorda/Desktop/Capstone Obesity/2025_CHRD.xlsx", sheet = "Select Measure Data", skip = 1)

data_2025 <- data_2025 |> 
  janitor::clean_names() 

food_index_2025 <- food_index_2025 |> 
  janitor::clean_names() 

data_2025 <- merge(data_2025, food_index_2025, by = "fips") 
data_2025 <- data_2025[, !grepl("percent_ci", names(data_2025))]
data_2025 <- data_2025[, !grepl("national_z", names(data_2025))]
data_2025 <- data_2025[, !grepl("injury_death", names(data_2025))]
data_2025 <- data_2025[, !grepl("percentile", names(data_2025))]
data_2025 <- data_2025[, !grepl("_white", names(data_2025))]
data_2025 <- data_2025[, !grepl("_hispanic", names(data_2025))]
data_2025 <- data_2025[, !grepl("_native", names(data_2025))]
data_2025 <- data_2025[, !grepl("_aian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_asian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_black", names(data_2025))]
data_2025 <- data_2025[, !grepl("dentist", names(data_2025))]
data_2025 <- data_2025[, !grepl("unreliable", names(data_2025))]
data_2025 <- data_2025[, !grepl("deaths", names(data_2025))]

drop_cols <- c("child_mortality_rate","hiv_prevalence_rate", " suicide_rate_age_adjusted", "crude_rate","teen_birth_rate","chlamydia_rate","number_chlamydia_cases", "drug_overdose_mortality_rate","other_primary_care_provider_rate", "other_primary_care_provider_ratio","percent_census_participation", "average_grade_performance_252", " average_grade_performance_246","cohort_size","gender_pay_gap","segregation_index_287","homicide_rate","motor_vehicle_mortality_rate", "percent_disconnected_youth","firearm_fatalities_rate","percent_lacking_support", "percent_low_birth_weight","number_primary_care_physicians","primary_care_physicians_ratio","primary_care_physicians_rate","number_mental_health_providers", "mental_health_provider_rate", "mental_health_provider_ratio", "visits_per_service_area_population", "presence_of_water_violation","average_daily_pm2_5", "state.y","county.y", "percent_with_annual_mammogram", "preventable_hospitalization_rate", "years_of_potential_life_lost_rate","number_firearm_fatalities","child_care_centers_per_1_000_children", "number_firearm_fatalities", "number_child_care_centers","spending_per_pupil","segregation_index_258","high_school_graduation_rate","average_grade_performance_246","disaster_declarations","weeks_in_moderate_or_greater_drought", "life_expectancy", "severe_housing_cost_burden", "percent_adults_with_diabetes","suicide_rate_age_adjusted","percent_fair_or_poor_health", "percent_voter_turnout", "percent_with_disability", "age_adjusted_death_rate", "percent_drive_alone_to_work", "income_ratio", "mens_median_earnings", "womens_median_earnings", "percent_with_access_to_exercise_opportunities", "infant_mortality_rate")

data_2025 <- data_2025[,!(names(data_2025) %in% drop_cols)]

data_2025 <- data_2025  |> 
  rename(
    county = county.x,
    state = state.x
  )

old_ct_fips <- c("09001", "09003", "09005", "09007", "09009", "09011", "09013", "09015")  

statewide_food_index <- data_2025 |> 
  filter(is.na(county) & !is.na(food_environment_index)) |> 
  select(state, food_environment_index) |> 
  rename(statewide_food_environment_index = food_environment_index)

data_2025 <- data_2025 |> 
  filter(!(fips %in% old_ct_fips)) |> 
  left_join(statewide_food_index, by = "state") |> 
  mutate(
    food_environment_index = if_else(
      is.na(food_environment_index) & !is.na(county),
      statewide_food_environment_index,
      food_environment_index),
      days_above_90f = if_else(days_above_90f > 365, 365, days_above_90f)) |> 
  select(-statewide_food_environment_index) |> 
  filter(!is.na(percent_adults_with_obesity))
```

## As physical inactivity increases obesity rates increases and as access to healthy food increases obesity rates decreases.

```{r scatterplot}
library(patchwork)

physical_obesity_plot <- data_2025 |>
  ggplot(aes(x = percent_physically_inactive, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "% Physically Inactive",
    y = "% Adults with Obesity"
  ) +
  theme_minimal()

food_obesity_plot <- data_2025 |>
  ggplot(aes(x = food_environment_index, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "Food Environment Index (0 = worst, 10 = best)",
    y = "% Adults with Obesity"
  ) +
  theme_minimal()

physical_obesity_plot + food_obesity_plot 
```

## Obesity rates are consistently higher in the Midwest and the South compared to the West Coast and the Northeast.

```{r obesity_map}
library(usmap)

q_brks <- floor(quantile(data_2025$percent_adults_with_obesity, probs = seq(0, 1, length.out = 5), na.rm = TRUE))

plot_usmap( data = data_2025, values = "percent_adults_with_obesity") +
  scale_fill_gradient(low = "lightyellow", high = "darkred",
    breaks = q_brks,
    labels = round(q_brks, 1),
    name = "Obesity Rate (%)",
    na.value = "grey90"
  ) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "bold")
  )
```

## Clustering U.S. Counties by Obesity, Food Access, and Inactivity

```{r dendro, include = FALSE}
library(ggdendro)

cluster_data <- data_2025 |>
  select(fips,
         state,
         county,
         percent_adults_with_obesity,
         percent_physically_inactive,
         food_environment_index) |> 
  drop_na()

std_data_features <- cluster_data |> 
  select(percent_adults_with_obesity, percent_physically_inactive, food_environment_index) |> 
  scale()

county_dist <- dist(std_data_features, method = "euclidean")
hc_complete <- hclust(county_dist, method = "complete")

ggdendrogram(hc_complete, labels = FALSE, leaf_labels = FALSE, theme_dendro = FALSE) +    labs(title = "Complete Linkage", y = "Dissimilarity") +
  theme_minimal() +
  theme(
      axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      plot.title = element_text(size = 12, face = "bold"),
      panel.grid = element_blank()
  )
```

```{r clustering}
library(knitr)

data_with_clusters <- cluster_data |> 
  mutate(cluster = as.factor(cutree(hc_complete, k = 4)))

plot_usmap(data = data_with_clusters, values = "cluster") +
  scale_fill_manual(
    values = ggthemes::colorblind_pal()(4),
    name = "Cluster",
    na.value = "gray90"
  ) +
  theme(legend.position = "left",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.title = element_text(face = "bold"))

data_2025 <- data_2025 |> 
  left_join(data_with_clusters |> select(fips, cluster), by = "fips")
```

## Key Attributes of Each Cluster

```{r}
data_2025 |> 
  filter(!is.na(cluster)) |> 
  group_by(cluster) |> 
  summarize(
    `Cluster Size` = n(),
    `Adults with Obesity (%)` = mean(percent_adults_with_obesity, na.rm = TRUE),
    `Food Environment Index` = mean(food_environment_index, na.rm = TRUE),
    `Physically Inactive (%)` = mean(percent_physically_inactive, na.rm = TRUE),
    `Access To Parks (%)` = mean(percent_with_access_to_parks, na.rm = TRUE),
    `Days Above 90F` = mean(days_above_90f, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  rename(`Cluster` = cluster) |> 
  kable(digits = 2)
```

From the clustering, we see that most of the U.S. has low park access. We thought that improving park access could help reduce inactivity. But in hotter regions like Clusters 1 and 3 in the Southeast, indoor facilities should be prioritized because of the extreme heat. This is one recommendation, and more will come later.

## Surprisingly, the Food Environment Index does not rank among the top predictors of county-level obesity rates.

```{r vip}
library(caret)
library(vip)

set.seed(10)
data_2025 <- data_2025 |> na.omit()

tree <- train(percent_adults_with_obesity ~ . - fips - state - county - cluster,
              method = "rpart", tuneLength = 20,
              trControl = trainControl(method = "cv", number = 10),
              data = data_2025)

top_vi <- head(vi(tree), 10)

pretty_labels <- c(
  "percent_physically_inactive" = "Physically Inactive",
  "median_household_income" = "Median Household Income",
  "percent_adults_reporting_currently_smoking" = "Smoking",
  "percent_excessive_drinking" = "Excessive Drinking",
  "percent_below_18_years_of_age" = "Population Below 18 Years Old",
  "percent_insufficient_sleep" = "Insufficient Sleep",
  "school_funding_adequacy" = "School Funding Adequacy",
  "percent_households_with_severe_cost_burden" = "Severe Cost Burden",
  "percent_uninsured_adults" = "Uninsured Adults",
  "traffic_volume" = "Traffic Volume",
  "food_environment_index" = "Food Environment Index"
)

top_vi$Variable <- ifelse(
  top_vi$Variable %in% names(pretty_labels),
  pretty_labels[top_vi$Variable],
  top_vi$Variable
)

vip(top_vi, aesthetics = list(fill = "steelblue")) +
  labs(y = "Variable Importance (Total Reduction in RSS)")

#TODO: how to decide how many variables to use in our beta regression because I just arbitarily chose 10 variables for this plot
```

## Beta Models

```{r beta_models}
library(betareg)
library(margins)


# lm_simple <- glm(percent_adults_with_obesity ~ food_environment_index + percent_physically_inactive, data = data_2025)
# summary(lm_simple)
# plot(lm_simple, which = 1:4)

data_2025 <- data_2025 |>
  mutate(obesity_prop = percent_adults_with_obesity / 100)

beta_simple <- betareg(obesity_prop ~ food_environment_index + percent_physically_inactive, data = data_2025, link = "logit")
beta_full <- betareg(obesity_prop ~ percent_physically_inactive + median_household_income + 
                      percent_adults_reporting_currently_smoking + percent_excessive_drinking + 
                      percent_below_18_years_of_age + percent_insufficient_sleep + 
                      school_funding_adequacy + percent_households_with_severe_cost_burden + 
                      percent_uninsured_adults + traffic_volume + food_environment_index, link = "logit",
                    data = data_2025)

me_simple <- margins(beta_simple)
me_full <- margins(beta_full)

df_simple <- summary(me_simple)
df_simple$model <- "Simple Model"
df_full <- summary(me_full)
df_full$model <- "Full Model"

margin_df <- rbind(df_simple, df_full)

margin_df <- margin_df |> 
  mutate(
    AME_percent = AME * 100,
    Lower = lower * 100,
    Upper = upper * 100,
    signif = ifelse(p < 0.01, 1, 0),
    factor = ifelse(factor %in% names(pretty_labels), 
                         pretty_labels[factor], 
                         factor)
  )

ggplot(margin_df, aes(x = reorder(factor, AME_percent), y = AME_percent, color = model)) +
  geom_point(position = position_dodge(width = 0.5), size = 1) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), 
                width = 0.2, position = position_dodge(width = 0.6)) +
  coord_flip() +
  labs(
    x = "Predictor",
    y = "Effect on Obesity Rate (%)",
    color = "Model"
  ) 

# margin_df |> 
#   select(factor, model, AME_percent, Lower, Upper, p) |>
#   kable(col.names = c("Predictor", "Model", "Effect (%)", "Lower CI", "Upper CI", "P-value"), digits = 2)
```

```{r beta_simple_diagnostics, eval = FALSE}
plot(beta_simple, which = 1:4) # does not look like it pass the tests?

```

```{r beta_full_diagnostics, eval = FALSE}
plot(beta_full, which = 1:4) # multicollinearity issues

predictors <- data_2025 |> 
  select(names(pretty_labels))

library(corrplot)
cor_matrix <- cor(predictors, use = "complete.obs")
rownames(cor_matrix) <- pretty_labels[rownames(cor_matrix)]
colnames(cor_matrix) <- pretty_labels[colnames(cor_matrix)]

corrplot(cor_matrix, 
         method = "color", 
         type = "upper", 
         tl.cex = 0.6, 
         tl.col = "black",
         addCoef.col = "black", 
         number.cex = .5)

#TODO: Identify highly correlated pairs, to decide  which variables to remove consider keeping the one that’s more interpretable or more strongly linked to the outcome idk ahhh
# i think removing insuffiencet sleep and income fixed it??
```

## Recommendations

-   Prioritize the young population by targeting environments that shape long-term health behaviors early in life.

-   Enhance sidewalk infrastructure across all clusters to promote walkability, encourage physical activity, and support safe routes for children and families.

-   Invest in school funding to strengthen the local food environment index and promote healthy habits—such as walking to school and accessing nutritious, affordable meals.
