---
title: "Mapping Obesity: A County-Level Analysis of Lifestyle and Food Environment Factors"
author: Helen Peng (Carnegie Mellon University), Jordan Faustin (Howard University), Anvith Thumma (Arizona State University)
format: 
  revealjs:
    theme: simple
    slide-number: true
    smaller: true
editor: visual
---
## Introduction 

Does physical inactivity level and access to healthy food affect obesity rate at the county level?

## Agenda

- **Significance**
- **Data Overview**
- **Scatter Analysis**
- **Percentage Distribution**
- **Predictor Analysis**
- **Regression Model**
- **Model Comparison**
- **Recommendations**
- **Appendix**

## Why This Question is Important

-   **Public Health Implications:** Obesity is a major contributor to rising healthcare costs and chronic diseases such as diabetes, heart disease, and cancer.

-   **Social Factors of Health:** Physical Inactivity and food access are modifiable factors linked to obesity. In turn, studying their impact highlights how our environment and infrastructure, like parks, grocery stores, and schools, can shape our health outcomes.

-   **Policy and Community Planning:** New information made through findings can inform policies like urban planning, subsidizing healthy food retailers in under-served areas, school/workplace wellness programs. 

## Data Overview

-   **Observations:** Data at the county level across all 50 U.S. states in 2025

-   **Variables of Interest:**

    -   \% Adults with Obesity

    -   \% Physically Inactive

    -   Food Environment Index (1-10) - Higher is Better

    -   State and County Identifiers
    
    - Days above 90°F
    
    _ Access to Parks

-   **Handling Missing Data:**

    -   If data on Food Environment Index or Physical Inactivity was missing for a county we used the state average.

    -   Seven counties that did not have data for obesity rate, our response variable and so, we dropped those counties to do a complete case analysis.

```{r setup}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center",
               echo=FALSE, tidy=TRUE)
```

```{r cleaning}
library(readxl)
library(tidyverse)
theme_set(theme_bw())

data_2025 <- read_excel("C:/Users/jorda/Desktop/Capstone Obesity/2025_CHRD.xlsx", sheet = "Additional Measure Data", skip = 1)

food_index_2025 <- read_excel("C:/Users/jorda/Desktop/Capstone Obesity/2025_CHRD.xlsx", sheet = "Select Measure Data", skip = 1)

data_2025 <- data_2025 |> 
  janitor::clean_names() 

food_index_2025 <- food_index_2025 |> 
  janitor::clean_names() 

data_2025 <- merge(data_2025, food_index_2025, by = "fips") 
data_2025 <- data_2025[, !grepl("percent_ci", names(data_2025))]
data_2025 <- data_2025[, !grepl("national_z", names(data_2025))]
data_2025 <- data_2025[, !grepl("injury_death", names(data_2025))]
data_2025 <- data_2025[, !grepl("percentile", names(data_2025))]
data_2025 <- data_2025[, !grepl("_white", names(data_2025))]
data_2025 <- data_2025[, !grepl("_hispanic", names(data_2025))]
data_2025 <- data_2025[, !grepl("_native", names(data_2025))]
data_2025 <- data_2025[, !grepl("_aian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_asian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_black", names(data_2025))]
data_2025 <- data_2025[, !grepl("dentist", names(data_2025))]
data_2025 <- data_2025[, !grepl("unreliable", names(data_2025))]
data_2025 <- data_2025[, !grepl("deaths", names(data_2025))]

drop_cols <- c("child_mortality_rate","hiv_prevalence_rate", " suicide_rate_age_adjusted", "crude_rate","teen_birth_rate","chlamydia_rate","number_chlamydia_cases", "drug_overdose_mortality_rate","other_primary_care_provider_rate", "other_primary_care_provider_ratio","percent_census_participation", "average_grade_performance_252", " average_grade_performance_246","cohort_size","gender_pay_gap","segregation_index_287","homicide_rate","motor_vehicle_mortality_rate", "percent_disconnected_youth","firearm_fatalities_rate","percent_lacking_support", "percent_low_birth_weight","number_primary_care_physicians","primary_care_physicians_ratio","primary_care_physicians_rate","number_mental_health_providers", "mental_health_provider_rate", "mental_health_provider_ratio", "visits_per_service_area_population", "presence_of_water_violation","average_daily_pm2_5", "state.y","county.y", "percent_with_annual_mammogram", "preventable_hospitalization_rate", "years_of_potential_life_lost_rate","number_firearm_fatalities","child_care_centers_per_1_000_children", "number_firearm_fatalities", "number_child_care_centers","spending_per_pupil","segregation_index_258","high_school_graduation_rate","average_grade_performance_246","disaster_declarations","weeks_in_moderate_or_greater_drought", "life_expectancy", "severe_housing_cost_burden", "percent_adults_with_diabetes","suicide_rate_age_adjusted","percent_fair_or_poor_health", "percent_voter_turnout", "percent_with_disability", "age_adjusted_death_rate", "percent_drive_alone_to_work", "income_ratio", "mens_median_earnings", "womens_median_earnings", "infant_mortality_rate","percent_uninsured_children","percent_uninsured_adults","percent_with_access_to_exercise_opportunities")

data_2025 <- data_2025[,!(names(data_2025) %in% drop_cols)]

data_2025 <- data_2025  |> 
  rename(
    county = county.x,
    state = state.x
  )

old_ct_fips <- c("09001", "09003", "09005", "09007", "09009", "09011", "09013", "09015")  

statewide_food_index <- data_2025 |> 
  filter(is.na(county) & !is.na(food_environment_index)) |> 
  dplyr::select(state, food_environment_index) |> 
  rename(statewide_food_environment_index = food_environment_index)

data_2025 <- data_2025 |> 
  filter(!(fips %in% old_ct_fips)) |> 
  left_join(statewide_food_index, by = "state") |> 
  mutate(
    food_environment_index = if_else(
      is.na(food_environment_index) & !is.na(county),
      statewide_food_environment_index,
      food_environment_index),
      days_above_90f = if_else(days_above_90f > 365, 365, days_above_90f)) |> 
  dplyr::select(-statewide_food_environment_index) |> 
  filter(!is.na(percent_adults_with_obesity))
```

## As physical inactivity increases obesity rates increases and as access to healthy food increases obesity rates decreases.

```{r scatterplot}
library(patchwork)

physical_obesity_plot <- data_2025 |>
  ggplot(aes(x = percent_physically_inactive, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "% Physically Inactive",
    y = "% Adults with Obesity"
  ) +
  theme_minimal()

food_obesity_plot <- data_2025 |>
  ggplot(aes(x = food_environment_index, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "Food Environment Index (0 = worst, 10 = best)",
    y = "% Adults with Obesity"
  ) +
  theme_minimal()

physical_obesity_plot + food_obesity_plot 
```

## Obesity rates are consistently higher in the Midwest and the South compared to the West Coast and the Northeast.

```{r obesity_map}
library(usmap)
library(viridis)
library(ggplot2)

# Define fixed limits based on your data range (18 to 53)
legend_min <- 18
legend_max <- 53

# Create quantile breaks (optional, but helps distribute colors)
q_brks <- floor(quantile(
  data_2025$percent_adults_with_obesity,
  probs = seq(0, 1, length.out = 5),
  na.rm = TRUE
))

# Plot map with FIXED legend scaling
plot_usmap(data = data_2025, values = "percent_adults_with_obesity", color = NA) +
  scale_fill_viridis_c(
    option = "cividis",     # Blue-yellow color scale (colorblind-friendly)
    direction = -1,         # Reverse to emphasize high values
    breaks = q_brks,        # Quantile breaks
    labels = round(q_brks, 1),
    name = "Obesity Rate (%)",
    na.value = "grey90",
    limits = c(legend_min, legend_max)  # FIXED SCALE (18-53)
  ) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "bold")
  )
```

## Surprisingly, the Food Environment Index does not rank among the top predictors of county-level obesity rates.

```{r vip}
# Enhanced CART-algorithm regression tree with 10-fold CV
library(caret)
library(vip)
library(rpart)       # For direct access to rpart functions
library(ggplot2)     # Enhanced plotting
library(dplyr)       # For data manipulation

set.seed(10)

# Improved data handling - consider alternative to na.omit()
data_2025_clean <- data_2025 %>% 
  select(-fips, -state, -county) %>%  # Remove ID columns upfront
  na.omit() %>%                      # Still using na.omit but more controlled
  mutate(across(where(is.character), as.factor))  # Convert chars to factors

# Define more controlled tuning grid
tune_grid <- expand.grid(
  cp = seq(0.001, 0.1, length.out = 20)  # More focused cp range

# Enhanced train control with additional options
ctrl <- trainControl(
  method = "cv",
  number = 10,
  savePredictions = "final",
  selectionFunction = "best"
)

# Train model with additional parameters
tree <- train(
  percent_adults_with_obesity ~ .,
  data = data_2025_clean,
  method = "rpart",
  tuneGrid = tune_grid,
  trControl = ctrl,
  metric = "RMSE",          # Explicit optimization metric
  maximize = FALSE,         # For RMSE we want to minimize
  control = rpart.control(
    minsplit = 20,          # Minimum observations to split
    minbucket = 7,          # Minimum observations in terminal node
    maxdepth = 10           # Maximum tree depth
  )
)

# Model evaluation
print(tree)
ggplot(tree) + ggtitle("Model Tuning Performance")
plot(tree$finalModel)
text(tree$finalModel, cex = 0.8, use.n = TRUE)

# Variable importance with better handling
pretty_labels <- c(
  "percent_physically_inactive" = "Physical Inactivity",
  "median_household_income" = "Median Income",
  "percent_adults_reporting_currently_smoking" = "Smoking Prevalence",
  "percent_excessive_drinking" = "Excessive Drinking",
  "percent_below_18_years_of_age" = "Under 18 Population",
  "percent_insufficient_sleep" = "Insufficient Sleep",
  "school_funding_adequacy" = "School Funding",
  "percent_households_with_severe_cost_burden" = "Housing Cost Burden",
  "percent_uninsured_adults" = "Uninsured Adults",
  "traffic_volume" = "Traffic Volume",
  "food_environment_index" = "Food Environment"
)

# Enhanced variable importance visualization
var_imp <- vip::vi(tree) %>%
  mutate(Variable = case_when(
    Variable %in% names(pretty_labels) ~ pretty_labels[Variable],
    TRUE ~ Variable
  )) %>%
  arrange(desc(Importance)) %>%
  head(15)

ggplot(var_imp, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Variable Importance (Total Reduction in RSS)",
    title = "Top Predictors of Adult Obesity Rates",
    subtitle = "Based on CART Regression Tree Analysis"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

## Linear Regression Model

```{r bivariate_model}
library(MASS)

data_2025 <- data_2025 |> 
  mutate(
    obesity_prop = percent_adults_with_obesity / 100,
    logit_obesity = log(obesity_prop / (1 - obesity_prop))
  )

lm_simple <- lm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
    data = data_2025)

summary(lm_simple)
# plot(lm_simple) -> Moved to appendix

huber_model <- rlm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
    data = data_2025)

summary(huber_model)
# Need to justify this over lm_simple
```

```{r multivariate_model}

```

## Model Comparsion

## Recommendations

-   Prioritize the young population by targeting environments that shape long-term health behaviors early in life.

-   Enhance sidewalk infrastructure across all clusters to promote walkability, encourage physical activity, and support safe routes for children and families.

-   Invest in school funding to strengthen the local food environment index and promote healthy habits—such as walking to school and accessing nutritious, affordable meals.

## Appendix

-   Correlation Matrix
-   Diagnostic Plots for Linear Regression Models
-   
