---
title: "Does physical inactivity level and access to healthy food affect obesity rate at the county level?"
author: Helen Peng, Anvith Thumma, and Jordan Faustin
format: 
  revealjs:
    theme: simple
    slide-number: true
    smaller: true
editor: visual
---

## Why This Question is Important

-   **Public Health Crisis:** Obesity is a major contributor to chronic diseases (diabetes, heart disease, cancer) and rising healthcare costs.

-   **Disparities in Obesity:** Obesity rates vary widely by region, income, and race/ethnicity. Identifying structural barriers (like food access and inactivity) can inform equitable policies.

-   **Rise in Weight Loss Drugs:** Medications like Ozempic and Wegovy are gaining popularity as quick fixes, but they’re expensive and don’t address root causes like environment or behavior.

## Data Overview

-   **Observations:** Data at the county level across all 50 U.S. states in 2025

-   **Variables of Interest:**

    -   \% Adults with Obesity

    -   \% Physically Inactive

    -   Food Environment Index (1-10) - Higher is Better

    -   State and County Identifiers

-   **Handling Missing Data:**

    -   If data on Food Environment Index or Physical Inactivity was missing for a county we used the state average.

    -   Seven counties that did not have data for obesity rate, our response variable and so, we dropped those counties to do a complete case analysis.

```{r setup}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center",
               echo=FALSE, tidy=TRUE)
```

```{r cleaning}
library(readxl)
library(tidyverse)
theme_set(theme_bw())

data_2025 <- read_excel('/Users/anviththumma/Desktop/UHG-Summer-2025/Capstone/UHG/CMUobesitycapstone/2025_CHRD.xlsx', sheet = "Additional Measure Data", skip = 1)

food_index_2025 <- read_excel('/Users/anviththumma/Desktop/UHG-Summer-2025/Capstone/UHG/CMUobesitycapstone/2025_CHRD.xlsx', sheet = "Select Measure Data", skip = 1)

data_2025 <- data_2025 |> 
  janitor::clean_names() 

food_index_2025 <- food_index_2025 |> 
  janitor::clean_names() 

data_2025 <- merge(data_2025, food_index_2025, by = "fips") 
data_2025 <- data_2025[, !grepl("percent_ci", names(data_2025))]
data_2025 <- data_2025[, !grepl("national_z", names(data_2025))]
data_2025 <- data_2025[, !grepl("injury_death", names(data_2025))]
data_2025 <- data_2025[, !grepl("percentile", names(data_2025))]
data_2025 <- data_2025[, !grepl("_white", names(data_2025))]
data_2025 <- data_2025[, !grepl("_hispanic", names(data_2025))]
data_2025 <- data_2025[, !grepl("_native", names(data_2025))]
data_2025 <- data_2025[, !grepl("_aian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_asian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_black", names(data_2025))]
data_2025 <- data_2025[, !grepl("dentist", names(data_2025))]
data_2025 <- data_2025[, !grepl("unreliable", names(data_2025))]
data_2025 <- data_2025[, !grepl("deaths", names(data_2025))]

drop_cols <- c("child_mortality_rate","hiv_prevalence_rate", " suicide_rate_age_adjusted", "crude_rate","teen_birth_rate","chlamydia_rate","number_chlamydia_cases", "drug_overdose_mortality_rate","other_primary_care_provider_rate", "other_primary_care_provider_ratio","percent_census_participation", "average_grade_performance_252", " average_grade_performance_246","cohort_size","gender_pay_gap","segregation_index_287","homicide_rate","motor_vehicle_mortality_rate", "percent_disconnected_youth","firearm_fatalities_rate","percent_lacking_support", "percent_low_birth_weight","number_primary_care_physicians","primary_care_physicians_ratio","primary_care_physicians_rate","number_mental_health_providers", "mental_health_provider_rate", "mental_health_provider_ratio", "visits_per_service_area_population", "presence_of_water_violation","average_daily_pm2_5", "state.y","county.y", "percent_with_annual_mammogram", "preventable_hospitalization_rate", "years_of_potential_life_lost_rate","number_firearm_fatalities","child_care_centers_per_1_000_children", "number_firearm_fatalities", "number_child_care_centers","spending_per_pupil","segregation_index_258","high_school_graduation_rate","average_grade_performance_246","disaster_declarations","weeks_in_moderate_or_greater_drought", "life_expectancy", "severe_housing_cost_burden", "percent_adults_with_diabetes","suicide_rate_age_adjusted","percent_fair_or_poor_health", "percent_voter_turnout", "percent_with_disability", "age_adjusted_death_rate", "percent_drive_alone_to_work", "income_ratio", "mens_median_earnings", "womens_median_earnings", "percent_with_access_to_exercise_opportunities", "infant_mortality_rate")

data_2025 <- data_2025[,!(names(data_2025) %in% drop_cols)]

data_2025 <- data_2025  |> 
  rename(
    county = county.x,
    state = state.x
  )

old_ct_fips <- c("09001", "09003", "09005", "09007", "09009", "09011", "09013", "09015")  

statewide_food_index <- data_2025 |> 
  filter(is.na(county) & !is.na(food_environment_index)) |> 
  dplyr::select(state, food_environment_index) |> 
  rename(statewide_food_environment_index = food_environment_index)

data_2025 <- data_2025 |> 
  filter(!(fips %in% old_ct_fips)) |> 
  left_join(statewide_food_index, by = "state") |> 
  mutate(
    food_environment_index = if_else(
      is.na(food_environment_index) & !is.na(county),
      statewide_food_environment_index,
      food_environment_index),
      days_above_90f = if_else(days_above_90f > 365, 365, days_above_90f)) |> 
  dplyr::select(-statewide_food_environment_index) |> 
  filter(!is.na(percent_adults_with_obesity))
```

## As physical inactivity increases obesity rates increases and as access to healthy food increases obesity rates decreases.

```{r scatterplot}
library(patchwork)

physical_obesity_plot <- data_2025 |>
  ggplot(aes(x = percent_physically_inactive, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "% Physically Inactive",
    y = "% Adults with Obesity"
  ) +
  theme_minimal()

food_obesity_plot <- data_2025 |>
  ggplot(aes(x = food_environment_index, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "Food Environment Index (0 = worst, 10 = best)",
    y = "% Adults with Obesity"
  ) +
  theme_minimal()

physical_obesity_plot + food_obesity_plot 
```

## Obesity rates are consistently higher in the Midwest and the South compared to the West Coast and the Northeast.

```{r obesity_map}
library(usmap)
library(viridis)
library(ggplot2)

# Create quantile breaks
q_brks <- floor(quantile(data_2025$percent_adults_with_obesity, probs = seq(0, 1, length.out = 5), na.rm = TRUE))

# Plot map
plot_usmap(data = data_2025, values = "percent_adults_with_obesity", color = NA) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1, 
    breaks = q_brks,
    labels = round(q_brks, 1),
    name = "Obesity Rate (%)",
    na.value = "grey90"
  ) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "bold"),
  ) 
```

## Clustering U.S. Counties by Obesity, Food Access, and Inactivity

```{r dendro, include = FALSE}
library(ggdendro)

cluster_data <- data_2025 |>
  dplyr::select(fips,
         state,
         county,
         percent_adults_with_obesity,
         percent_physically_inactive,
         food_environment_index) |> 
  drop_na()

std_data_features <- cluster_data |> 
  dplyr::select(percent_adults_with_obesity, percent_physically_inactive, food_environment_index) |> 
  scale()

county_dist <- dist(std_data_features, method = "euclidean")
hc_complete <- hclust(county_dist, method = "complete")

ggdendrogram(hc_complete, labels = FALSE, leaf_labels = FALSE, theme_dendro = FALSE) +    labs(title = "Complete Linkage", y = "Dissimilarity") +
  theme_minimal() +
  theme(
      axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      plot.title = element_text(size = 12, face = "bold"),
      panel.grid = element_blank()
  )
```

```{r clustering}
library(knitr)

data_with_clusters <- cluster_data |> 
  mutate(cluster = as.factor(cutree(hc_complete, k = 4)))

plot_usmap(data = data_with_clusters, values = "cluster") +
  scale_fill_manual(
    values = ggthemes::colorblind_pal()(4),
    name = "Cluster",
    na.value = "gray90"
  ) +
  theme(legend.position = "left",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.title = element_text(face = "bold"))

data_2025 <- data_2025 |> 
  left_join(data_with_clusters |> dplyr::select(fips, cluster), by = "fips")
```

## Key Attributes of Each Cluster

```{r}
data_2025 |> 
  filter(!is.na(cluster)) |> 
  group_by(cluster) |> 
  summarize(
    `Cluster Size` = n(),
    `Adults with Obesity (%)` = mean(percent_adults_with_obesity, na.rm = TRUE),
    `Food Environment Index` = mean(food_environment_index, na.rm = TRUE),
    `Physically Inactive (%)` = mean(percent_physically_inactive, na.rm = TRUE),
    `Access To Parks (%)` = mean(percent_with_access_to_parks, na.rm = TRUE),
    `Days Above 90F` = mean(days_above_90f, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  rename(`Cluster` = cluster) |> 
  kable(digits = 2)
```

From the clustering, we see that most of the U.S. has low park access. We thought that improving park access could help reduce inactivity. But in hotter regions like Clusters 1 and 3 in the Southeast, indoor facilities should be prioritized because of the extreme heat. This is one recommendation, and more will come later.

## Surprisingly, the Food Environment Index does not rank among the top predictors of county-level obesity rates.

```{r vip}
library(caret)
library(vip)

set.seed(10)
data_2025 <- data_2025 |> na.omit()

tree <- train(percent_adults_with_obesity ~ . - fips - state - county - cluster,
              method = "rpart", tuneLength = 20,
              trControl = trainControl(method = "cv", number = 10),
              data = data_2025)

top_vi <- head(vi(tree), 10)

pretty_labels <- c(
  "percent_physically_inactive" = "Physically Inactive",
  "median_household_income" = "Median Household Income",
  "percent_adults_reporting_currently_smoking" = "Smoking",
  "percent_excessive_drinking" = "Excessive Drinking",
  "percent_below_18_years_of_age" = "Population Below 18 Years Old",
  "percent_insufficient_sleep" = "Insufficient Sleep",
  "school_funding_adequacy" = "School Funding Adequacy",
  "percent_households_with_severe_cost_burden" = "Severe Cost Burden",
  "percent_uninsured_adults" = "Uninsured Adults",
  "traffic_volume" = "Traffic Volume",
  "food_environment_index" = "Food Environment Index"
)

top_vi$Variable <- ifelse(
  top_vi$Variable %in% names(pretty_labels),
  pretty_labels[top_vi$Variable],
  top_vi$Variable
)

vip(top_vi, aesthetics = list(fill = "steelblue")) +
  labs(y = "Variable Importance (Total Reduction in RSS)")

#TODO: how to decide how many variables to use in our beta regression because I just arbitarily chose 10 variables for this plot
```

## Linear Regression Model (Obesity)

```{r linear_models}
library(MASS)

data_2025 <- data_2025 |> 
  mutate(
    obesity_prop = percent_adults_with_obesity / 100,
    logit_obesity = log(obesity_prop / (1 - obesity_prop))
  )

lm_simple <- lm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
    data = data_2025)

summary(lm_simple)
# plot(lm_simple) -> Moved to appendix

huber_model <- rlm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
    data = data_2025)

summary(huber_model)
# Need to justify this over lm_simple
```

## Food Environment Index appears to be a confounding variable

```{r}
# Clean data
set.seed(10)
data_2025 <- data_2025 |> na.omit()

tree_fei <- train(food_environment_index ~ . - fips - state - county - cluster,
              method = "rpart", tuneLength = 20,
              trControl = trainControl(method = "cv", number = 10),
              data = data_2025)

# Top 10 predictors of FEI
top_vi_fei <- head(vi(tree_fei), 10)


top_vi_fei$Variable <- ifelse(
  top_vi_fei$Variable %in% names(pretty_labels_fei),
  pretty_labels_fei[top_vi_fei$Variable],
  top_vi_fei$Variable
)

vip(top_vi_fei, aesthetics = list(fill = "darkgreen")) +
  labs(y = "Variable Importance (Total Reduction in RSS)")

```

```{r, eval=FALSE}
summary(lm(logit_obesity ~ food_environment_index, data = data_2025))

summary(lm(logit_obesity ~ food_environment_index + percent_limited_access_to_healthy_foods + percent_food_insecure, data = data_2025))

summary(lm(logit_obesity ~ percent_physically_inactive + percent_limited_access_to_healthy_foods + percent_food_insecure, data = data_2025))

```

### Summary

During our exploratory data analysis (EDA), we observed that the **Food Environment Index (FEI)** had an inverse relationship with obesity — suggesting that counties with better food environments tend to have lower obesity rates. However, when we modeled this relationship using linear regression, the results became inconsistent. In one model, FEI appeared to be a strong negative predictor of obesity, but once we added its key components — **percent food insecure** and **percent with limited access to healthy foods** — FEI became statistically insignificant. This indicates that FEI is a **confounded composite variable**, meaning its apparent effect on obesity is actually driven by the variables that make up the index. Because of this, FEI is not a reliable standalone predictor of obesity.

To understand what drives FEI and explore its confounding nature, I built a decision tree to identify the top predictors of FEI. The two most influential variables were **% food insecure** and **% with limited access to healthy foods**. I then ran a new model including these two variables along with **% physically inactive**, a variable already shown to be a strong predictor of obesity. In this final model, **physical inactivity remained highly significant**, **limited access to healthy food was not significant**, and **food insecurity showed a significant but unexpected negative association** with obesity. These findings suggest that it is more effective to use **specific, interpretable variables** related to food access rather than relying on an aggregated index like FEI. This approach provides clearer insights into the factors influencing obesity at the county level and helps guide more targeted recommendations.

**Overall Key Takeaways:**

-   **Physical inactivity** is the strongest and most consistent predictor of obesity.

-   **% Food insecure** becomes a significant (negative) predictor once we control for physical inactivity and drop FEI.

-   **% Limited access to healthy food** is not significant in any specification we tested.

-   Relying on the FEI alone is misleading because its signal disappears once key components (and physical inactivity) are included.

## Recommendations

### Invest in Built Environments That Promote Daily Physical Activity

-   **TO DO:**

    -   Improve sidewalk infrastructure in neighborhoods with high inactivity and obesity rates.

    -   Add bike lanes, trails, and etc

    -   Zone for mixed-use neighborhoods to encourage walkable errands.

-   **Real-world example**:

    -   **Minneapolis, MN** implemented "Complete Streets" policies

        -   Street design policy that supports walking, biking, transit, and driving

        -   Includes sidewalks, bike lanes, crosswalks, lighting, and traffic calming

        -   Makes daily physical activity safer and more accessible

    -   **Safe Routes to School** programs in **California**

        -   National program to help students walk or bike safely to school

        -   Improves sidewalks, crosswalks, signage, and adds crossing guards

        -   Helps form lifelong physical activity habits starting in childhood

### Deploy Mobile Markets and Pop-Up Grocery Stands in High-Need Areas

-   **TO DO:**

    -   Use mobile grocery vans or pop-up farmers' markets in areas that lack supermarkets

    -   Accept **SNAP/WIC** benefits and keep prices affordable

        -   SNAP is essentially food stamps

        -   WIC is federal assistance for pregnant women, new mothers, and children under 5. Provides vouchers for nutritious foods, baby formula, and health care referrals and nutrition education

    -   Pair markets with cooking demonstrations or physical activity events

-   **Real-world example**:

    -   **Veggie Van (North Carolina)** brings produce into underserved neighborhoods and has partnered with local health departments

    -   **Fresh Moves Mobile Market (Chicago)** operates a bus converted into a traveling produce market for low-access communities
