---
title: "Mapping Obesity: A County-Level Analysis of Lifestyle and Food Environment Factors"
subtitle: "SURE 2025"
author:
  - "<strong>Helen Peng</strong><br><small>Carnegie Mellon University</small>"
  - "<strong>Jordan Faustin</strong><br><small>Howard University</small>"
  - "<strong>Anvith Thumma</strong><br><small>Arizona State University</small>"
format: 
  revealjs:
    theme: simple
    slide-number: true
    smaller: true
editor: visual
---

## Does physical inactivity level and access to healthy food affect obesity rate at the county level? {.center}

## Why This Question is Important

-   **Public Health Implications:** <br> Obesity is a major contributor to rising healthcare costs and chronic diseases such as diabetes, heart disease, and cancer.

-   **Social Factors of Health:** <br> Physical Inactivity and food access are modifiable factors linked to obesity. In turn, studying their impact highlights how our environment and infrastructure, like parks, grocery stores, and schools, can shape our health outcomes.

-   **Rise in Weight Loss Drugs:** Medications like Ozempic and Wegovy are gaining popularity as quick fixes, but they’re expensive and don’t address root causes like environment or behavior.

-   **Policy and Community Planning:** <br> New information made through findings can inform policies like urban planning, subsidizing healthy food retailers in under-served areas, school/workplace wellness programs.

## Agenda

-   **Data Overview**
-   **Initial Analysis**
-   **Understanding Direction**
-   **Modeling and Comparison**
-   **Discussion**
-   **Recommendations**
-   **Appendix**

## Data Overview

-   **Observations:**

    -   Data at the county level across all 50 U.S. states in 2025

-   **Variables of Interest:**

    -   \% Adults with Obesity

    -   \% Physically Inactive

    -   Food Environment Index (1-10)

    -   State and County Identifiers

-   **Handling Missing Data:**

    -   If data on Food Environment Index or Physical Inactivity was missing for a county we used the state average.

    -   Seven counties that did not have data for obesity rate, our response variable and so, we dropped those counties to do a complete case analysis.

```{r setup}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center",
               echo=FALSE, tidy=TRUE)
```

```{r cleaning}
library(readxl)
library(tidyverse)
theme_set(theme_bw())

data_2025 <- read_excel("C:/Users/helen/OneDrive/Documents/GitHub/obesity-capstone/Peng/2025_data.xlsx", sheet = "Additional Measure Data", skip = 1)

food_index_2025 <- read_excel("C:/Users/helen/OneDrive/Documents/GitHub/obesity-capstone/Peng/2025_data.xlsx", sheet = "Select Measure Data", skip = 1)

data_2025 <- data_2025 |> 
  janitor::clean_names() 

food_index_2025 <- food_index_2025 |> 
  janitor::clean_names() 

data_2025 <- merge(data_2025, food_index_2025, by = "fips") 
data_2025 <- data_2025[, !grepl("percent_ci", names(data_2025))]
data_2025 <- data_2025[, !grepl("national_z", names(data_2025))]
data_2025 <- data_2025[, !grepl("injury_death", names(data_2025))]
data_2025 <- data_2025[, !grepl("percentile", names(data_2025))]
data_2025 <- data_2025[, !grepl("_white", names(data_2025))]
data_2025 <- data_2025[, !grepl("_hispanic", names(data_2025))]
data_2025 <- data_2025[, !grepl("_native", names(data_2025))]
data_2025 <- data_2025[, !grepl("_aian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_asian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_black", names(data_2025))]
data_2025 <- data_2025[, !grepl("dentist", names(data_2025))]
data_2025 <- data_2025[, !grepl("unreliable", names(data_2025))]
data_2025 <- data_2025[, !grepl("deaths", names(data_2025))]

drop_cols <- c("child_mortality_rate","hiv_prevalence_rate", " suicide_rate_age_adjusted", "crude_rate","teen_birth_rate","chlamydia_rate","number_chlamydia_cases", "drug_overdose_mortality_rate","other_primary_care_provider_rate", "other_primary_care_provider_ratio","percent_census_participation", "average_grade_performance_252", " average_grade_performance_246","cohort_size","gender_pay_gap","segregation_index_287","homicide_rate","motor_vehicle_mortality_rate", "percent_disconnected_youth","firearm_fatalities_rate","percent_lacking_support", "percent_low_birth_weight","number_primary_care_physicians","primary_care_physicians_ratio","primary_care_physicians_rate","number_mental_health_providers", "mental_health_provider_rate", "mental_health_provider_ratio", "visits_per_service_area_population", "presence_of_water_violation","average_daily_pm2_5", "state.y","county.y", "percent_with_annual_mammogram", "preventable_hospitalization_rate", "years_of_potential_life_lost_rate","number_firearm_fatalities","child_care_centers_per_1_000_children", "number_firearm_fatalities", "number_child_care_centers","spending_per_pupil","segregation_index_258","high_school_graduation_rate","average_grade_performance_246","disaster_declarations","weeks_in_moderate_or_greater_drought", "life_expectancy", "severe_housing_cost_burden", "percent_adults_with_diabetes","suicide_rate_age_adjusted","percent_fair_or_poor_health", "percent_voter_turnout", "percent_with_disability", "age_adjusted_death_rate", "percent_drive_alone_to_work", "income_ratio", "mens_median_earnings", "womens_median_earnings", "infant_mortality_rate","percent_uninsured_children","percent_uninsured_adults","percent_with_access_to_exercise_opportunities")

data_2025 <- data_2025[,!(names(data_2025) %in% drop_cols)]

data_2025 <- data_2025  |> 
  rename(
    county = county.x,
    state = state.x
  )

old_ct_fips <- c("09001", "09003", "09005", "09007", "09009", "09011", "09013", "09015")  

statewide_food_index <- data_2025 |> 
  filter(is.na(county) & !is.na(food_environment_index)) |> 
  dplyr::select(state, food_environment_index) |> 
  rename(statewide_food_environment_index = food_environment_index)

data_2025 <- data_2025 |> 
  filter(!(fips %in% old_ct_fips)) |> 
  left_join(statewide_food_index, by = "state") |> 
  mutate(
    food_environment_index = if_else(
      is.na(food_environment_index) & !is.na(county),
      statewide_food_environment_index,
      food_environment_index),
      days_above_90f = if_else(days_above_90f > 365, 365, days_above_90f)) |> 
  dplyr::select(-statewide_food_environment_index) |> 
  filter(!is.na(percent_adults_with_obesity))
```

## As physical inactivity increases obesity rates increase as well, and as access to healthy food increase obesity rates start to decline

```{r scatterplot}
library(patchwork)

physical_obesity_plot <- data_2025 |>
  ggplot(aes(x = percent_physically_inactive, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "% Physically Inactive",
    y = "% Adults with Obesity"
  ) 

food_obesity_plot <- data_2025 |>
  ggplot(aes(x = food_environment_index, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "Food Environment Index (0 = worst, 10 = best)",
    y = "% Adults with Obesity"
  ) 

physical_obesity_plot + food_obesity_plot 
```

## Obesity rates are higher in the Midwest and South compared to the Northeast and West

```{r obesity_map}
library(usmap)
library(viridis)
library(ggplot2)

min_val <- floor(min(data_2025$percent_adults_with_obesity, na.rm = TRUE))
max_val <- ceiling(max(data_2025$percent_adults_with_obesity, na.rm = TRUE))

custom_breaks <- seq(min_val, max_val, length.out = 5)

plot_usmap(data = data_2025, values = "percent_adults_with_obesity", color = NA) +
  scale_fill_viridis_c(
    option = "inferno",
    direction = -1,
    limits = c(min_val, max_val),
    breaks = custom_breaks,
    labels = round(custom_breaks, 1),
    name = "Obesity Rate (%)",
    na.value = "grey90"
  ) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "bold")
  )

```

## Modeling

```{r}
library(broom)
library(MASS)

data_2025 <- data_2025 |> 
  mutate(
    obesity_prop = percent_adults_with_obesity / 100,
    logit_obesity = log(obesity_prop / (1 - obesity_prop))
  )

huber_model <- rlm(logit_obesity ~ food_environment_index + percent_physically_inactive, 
                   data = data_2025)

huber_model2 <- lm(logit_obesity ~ median_household_income + percent_limited_access_to_healthy_foods, 
                    data = data_2025)

coef1_df <- tidy(huber_model, conf.int = TRUE) %>% 
  filter(term != "(Intercept)") %>%
  mutate(
    model = "Model 1",
    OR = exp(estimate),
    OR_low = exp(conf.low),
    OR_high = exp(conf.high)
  )

coef2_df <- tidy(huber_model2, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    model = "Model 2",
    OR = exp(estimate),
    OR_low = exp(conf.low),
    OR_high = exp(conf.high)
  )

coef_df <- bind_rows(coef1_df, coef2_df) %>%
  mutate(
    term = recode(term,
                  "food_environment_index" = "Food Environment Index",
                  "percent_physically_inactive" = "Physically Inactive (%)",
                  "percent_limited_access_to_healthy_foods" = "Limited Access to Healthy Food (%)",
                  "median_household_income" = "Median Household Income")
  )

ggplot(coef_df, aes(x = OR, y = term, color = model)) +
  geom_point(size = 3, position = position_dodge(width = 0.6)) +
  geom_errorbarh(aes(xmin = OR_low, xmax = OR_high), height = 0.2, 
                 position = position_dodge(width = 0.6)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  labs(
    x = "Odds Ratio (Effect Size)",
    y = NULL,
    color = "Model"
  ) +
  theme_minimal()

```

```{r eval = FALSE}

## Food Environment Index is a composite score, shaped by other variables such as food insecurity and median household income. 
library(caret)
library(vip)
set.seed(10)
data_2025 <- data_2025 |> na.omit()

tree_fei <- train(food_environment_index ~ . - fips - state - county,
              method = "rpart", tuneLength = 20,
              trControl = trainControl(method = "cv", number = 10),
              data = data_2025)

top_vi_fei <- head(vi(tree_fei), 10)

pretty_labels_fei <- c(
  "percent_limited_access_to_healthy_foods" = "Limited Access to Healthy Foods (%)",
  "percent_food_insecure" = "Food Insecurity (%)",
  "median_household_income" = "Median Household Income",
  "percent_children_in_poverty" = "Children in Poverty (%)",
  "number_limited_access_to_healthy_foods" = "Number with Limited Access to Healthy Foods",
  "average_number_of_physically_unhealthy_days" = "Avg. Physically Unhealthy Days",
  "percent_frequent_physical_distress" = "Frequent Physical Distress (%)",
  "percent_physically_inactive" = "Physically Inactive (%)",
  "number_completed_high_school" = "Completed High School",
  "population_173" = "Population "
)

top_vi_fei$Variable <- ifelse(
  top_vi_fei$Variable %in% names(pretty_labels_fei),
  pretty_labels_fei[top_vi_fei$Variable],
  top_vi_fei$Variable
)

vip(top_vi_fei, aesthetics = list(fill = "darkgreen")) +
  labs(y = "Variable Importance (Total Reduction in RSS)")

```

```{r eval=FALSE}
summary(lm(logit_obesity ~ percent_food_insecure + percent_physically_inactive + percent_limited_access_to_healthy_foods, data = data_2025))

summary(lm(logit_obesity ~ percent_food_insecure + percent_physically_inactive, data = data_2025))

```

## Key Takeaways

-   **Physical Inactivity** is the most powerful and consistent predictor of county-level obesity rates.

-   **Food Insecurity (%)** becomes statistically significant *only* when controlling for physical inactivity and excluding the composite **Food Environment Index (FEI)**.

-   **Limited Access to Healthy Food (%)** does *not* significantly predict obesity in any model we tested.

-   **The Food Environment Index (FEI)** may be misleading as a standalone measure: once we account for its components and physical inactivity, its predictive value disappears.

## Recommendations

#### Prioritize Physical Activity Infrastructure

-   Building Walkable Communities (expanding sidewalks, bike lanes, and trails in high-obesity counties)
-   Funding Safe Routes to School programs to improve sidewalks, crosswalks, and safety patrols.

#### Improve Food Access with Targeted Interventions

-   Deploying Mobile Food Solutions such as produce vans (e.g., Veggie Van in NC) in food deserts. and hosting pop-up markets paired with nutrition workshops.

#### Redirecting Funding

-   Shifting resources from FEI-based programs to hyper-local initiatives (e.g., subsidized CSA boxes for low-income families).

## Possible Future Work

-   Establishing a count of parks and wellness centers per county

-   Mapping their respective distance and location

-   Checking for trends on when individuals are going to the fitness locations

-   Causal inference work like DAGs, Propensity scores and so on.

## Thank You {.center}

## Questions? {.center}

## Appendix

-   Correlation Matrix
-   Diagnostic Plots for Linear Regression Models

## Diagnostic Plots

```{r}
plot(lm_simple, which = 1:4)
```
