---
title: "Do physical inactivity level and access to healthy food affect obesity rate at the county level?"
author: "Helen Peng"
output: html_document
---

```{r cleaning}
library(readxl)
library(tidyverse)
theme_set(theme_bw())

data_2024 <- read_excel("C:/Users/helen/OneDrive/Documents/GitHub/obesity-capstone/Peng/2024_county_health_release_data_-_v1.xlsx", sheet = "Select Measure Data", skip = 1)

data_2025 <- read_excel("C:/Users/helen/OneDrive/Documents/GitHub/obesity-capstone/Peng/2025 County Health Rankings Data - v3.xlsx", sheet = "Select Measure Data", skip = 1)

data_2024 <- data_2024 |>
  select(FIPS, State, County, 80, 84, 86, 169) |>
  janitor::clean_names() |> 
  filter(!is.na(county))

data_2025 <- data_2025 |> 
  select(FIPS, State, County, 111, 147, 149, 215) |>
  janitor::clean_names() |> 
  filter(!is.na(county)) # could still use food environment index but we would need to merge the datasets
```

# Maps

```{r static_map_2024}
library(usmap)
plot_usmap(data = data_2024, values = "percent_adults_with_obesity", regions = "counties") +
  scale_fill_continuous(
    low = "lightyellow", high = "darkred", name = "Obesity Rate (%)"
  ) +
  theme(legend.position = "right")

data_2024[is.na(data_2024$percent_adults_with_obesity), ]
data_2024[data_2024$state == "Connecticut",]
```

```{r dynamic_map_2024}
library(tigris)
library(sf)
library(tmap)

options(tigris_use_cache = TRUE)

counties_2023 <- 
  counties(cb = TRUE, resolution = "5m", class = "sf") |> 
  mutate(fips = paste0(STATEFP, COUNTYFP))

data_2024 <- data_2024 |> 
  mutate(fips = sprintf("%05s", fips))

county_data_prechange <- counties_2023 |> 
  left_join(data_2024, by = "fips")

tmap_mode("view") 

tm_shape(county_data_prechange) +
  tm_polygons("percent_adults_with_obesity", 
              palette = "YlOrRd", 
              title = "Obesity Rate (%)",
              style = "quantile") +
  tm_layout(legend.outside = TRUE)
```

```{r mississippi}
library(patchwork)

# Statistics for Mississippi, the state with the highest obesity rate
ms_county <- data_2024 |> 
  filter(state == "Mississippi") |> 
  arrange(percent_adults_with_obesity)

ms_obesity_plot <- plot_usmap(data = ms_county, values = "percent_adults_with_obesity", regions = "counties", include = "MS") +
  scale_fill_continuous(
    low = "lightyellow", high = "darkred", name = "Obesity Rate (%)"
  ) +
  labs(title = "Average Adult Obesity Rate by County in Mississippi (2024)",
       subtitle = "Based on County Health Rankings Data") +
  theme(legend.position = "right")

ms_food_plot <- plot_usmap(data = ms_county, values = "food_environment_index", regions = "counties", include = "MS") +
  scale_fill_continuous(
    low = "lightyellow", high = "darkred", name = "Food Index Rating"
  ) +
  labs(title = "Average Food Index by County in Mississippi (2024)",
       subtitle = "Based on County Health Rankings Data") +
  theme(legend.position = "right")

ms_physical_plot <- plot_usmap(data = ms_county, values = "percent_physically_inactive", regions = "counties", include = "MS") +
  scale_fill_continuous(
    low = "lightyellow", high = "darkred", name = "Physically Inactive (%)"
  ) +
  labs(title = "Average Physically Inactive by County in Mississippi (2024)",
       subtitle = "Based on County Health Rankings Data") +
  theme(legend.position = "right")

combined_ms_plot <- 
  ms_obesity_plot + labs(title = NULL, subtitle = NULL) +
  ms_food_plot + labs(title = NULL, subtitle = NULL) +
  ms_physical_plot + labs(title = NULL, subtitle = NULL) +
  plot_annotation(
  title = "Statistics for Mississippi by County",
  subtitle = "Obesity, Food Access, and Physical Inactivity Rates"
)

combined_ms_plot
```

```{r dynamic_map_2025}
library(tidyverse)
library(tigris)
library(sf)
library(tmap)

options(tigris_use_cache = TRUE)

counties_sf <- counties(cb = TRUE, resolution = "5m", class = "sf") |>
  mutate(fips = paste0(STATEFP, COUNTYFP))

county_data <- counties_sf |> 
  left_join(data_2025, by = "fips")

tmap_mode("view") 

tm_shape(county_data) +
  tm_polygons(
    fill = "percent_adults_with_obesity",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "YlOrRd"
    ),
    fill.legend = tm_legend(title = "Obesity Rate (%)")
  ) +
  tm_layout(legend.outside = TRUE)
```

# Scatter plots

```{r scatterplot_2024}
library(patchwork)
food_obesity_plot <- data_2024 |> 
  filter(!is.na(food_environment_index), !is.na(percent_adults_with_obesity)) |> 
  ggplot(aes(x = food_environment_index, y = percent_adults_with_obesity)) +
  geom_point() +
  labs(x = "Food Environment Index (Higher = Better Access)",
       y = "Obesity Rate (%)") +
  geom_smooth(method = "lm")

physical_obesity_plot <- data_2024 |> 
  filter(!is.na(percent_physically_inactive), !is.na(percent_adults_with_obesity)) |> 
  ggplot(aes(x = percent_physically_inactive, y = percent_adults_with_obesity)) +
  geom_point() +
  labs(x = "Phyiscally Inactive Rate (%)",
       y = "Obesity Rate (%)") +
  geom_smooth(method = "lm")

combined_scatterplots <- food_obesity_plot + physical_obesity_plot +
  plot_annotation(title = "Food Enivronment Index and Physical Inactivity Rate vs. Obesity Rate")
combined_scatterplots
```

```{r pairplot_2024}
library(GGally)

pair_data <- data_2024 |> 
  select(percent_adults_with_obesity, 
         percent_physically_inactive, 
         food_environment_index) |> 
  rename(obesity_rate = percent_adults_with_obesity,
         food_access = food_environment_index,
         physically_inactive = percent_physically_inactive)
  drop_na()

ggpairs(pair_data,
        lower = list(continuous = wrap("smooth", method = "lm", se = FALSE, color = "blue")),
        diag = list(continuous = "densityDiag"), 
        upper = list(continuous = wrap("cor", size = 4))) +
  theme_minimal() +
  labs(title = "Pair Plot: Obesity, Physical Inactivity, and Food Environment Index")
```

# Hard Clustering (Hierarchical Clustering)

```{r dendro_2024}
library(ggdendro)

missing_data <- data_2024 |> 
  filter((is.na(percent_adults_with_obesity) |
          is.na(percent_physically_inactive) |
          is.na(food_environment_index)) &
         !is.na(fips)) |> 
  select(fips, state, county, percent_adults_with_obesity, percent_physically_inactive, food_environment_index)

data_2024 <- data_2024 |>
  select(fips,
         state,
         county,
         percent_adults_with_obesity,
         percent_physically_inactive,
         food_environment_index) |> 
  drop_na()

std_data_features <- data_2024 |> 
  select(percent_adults_with_obesity, percent_physically_inactive, food_environment_index) |> 
  scale()

county_dist <- dist(std_data_features, method = "euclidean")

hc_complete <- hclust(county_dist, method = "complete")
hc_single   <- hclust(county_dist, method = "single")
hc_average  <- hclust(county_dist, method = "average")

plot_dendro <- function(hc_obj, title_text) {
  ggdendrogram(hc_obj, labels = FALSE, leaf_labels = FALSE, theme_dendro = FALSE) +
    labs(title = title_text, y = "Dissimilarity") +
    theme_minimal() +
    theme(
      axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      plot.title = element_text(size = 12, face = "bold"),
      panel.grid = element_blank()
    )
}

dendro_complete <- plot_dendro(hc_complete, "Complete Linkage")
dendro_single   <- plot_dendro(hc_single, "Single Linkage")
dendro_average  <- plot_dendro(hc_average, "Average Linkage")

dendro_combined <- dendro_single + dendro_average + dendro_complete +
  plot_layout(ncol = 3)

dendro_combined
```

```{r clusters_2024}
fviz_cluster(
  list(data = std_data_features,
       cluster = cutree(hc_complete, k = 4)),
  geom = "point",
  ellipse = FALSE
) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

```{r map_clusters_2024}
data_with_clusters <- data_2024 |>   
  mutate(cluster = as.factor(cutree(hc_complete, k = 4)))

plot_usmap(data = data_with_clusters, values = "cluster", regions = "counties") +
  ggthemes::scale_fill_colorblind(name = "Cluster") +
  theme(legend.position = "right") 

```

```{r clusters_summary_2024}
library(knitr)

# Post-clustering analysis
cluster_summary <- data_with_clusters |>
  group_by(cluster) |>
  summarize(
    n_counties = n(),
    mean_physically_inactive = mean(percent_physically_inactive, na.rm = TRUE),
    mean_food_index = mean(food_environment_index, na.rm = TRUE),
    mean_obesity_rate = mean(percent_adults_with_obesity, na.rm = TRUE),
  )
print(cluster_summary)

cluster_summary |> 
  rename(
    `Cluster` = cluster,
    `Number of Counties` = n_counties,
    `Physically Inactive (%)` = mean_physically_inactive,
    `Food Environment Index` = mean_food_index,
    `Obesity Rate (%)` = mean_obesity_rate
  ) |> 
  kable(digits = 2, 
        caption = "Cluster Summary: Average Health and Food Metrics by Cluster")

```

Cluster 1: High inactivity, decent food access, high obesity\
Cluster 2: Very inactive, poor food access, highest obesity - Worst Case\
Cluster 3: Most active, best food access, lowest obesity - Best Case\
Cluster 4: Moderately active, poor food access, mid obesity

# Modeling

Note for Modeling: Response Variable is Continuous and Bounded. What models can we use to handle bounded variables? Beta regression??

## Simple Linear Regression

## Ridge Regression

## LASSO Regression

## Elastic Net Regression

## Generalized Additive Models

## Random Forest Models

## Gradient Boosting Models
