---
title: "Mapping Obesity: A County-Level Analysis of Lifestyle and Food Environment Factors"
author:
  - Helen Peng
  - Anvith Thumma
  - Jordon Faustin
date: "July 25, 2025"
toc: true
format:
  html:
    theme: cosmo
    html-math-method: katex
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
---

------------------------------------------------------------------------

## Introduction

Describe the problem and why it is important.

## Data

We used publicly available county-level data from the **2025 County Health Rankings Data**, provided by the University of Wisconsin Population Health Institute.
The dataset includes a wide range of health-related indicators across all U.S. counties, including demographics, socioeconomic, behavioral, and healthcare access variables.

Our main variables of interests were:

-   **Adults with Obesity (%):** Percentage of adults (18+) with a body mass index (BMI) ≥ 30 kg/m² (age-adjusted).

-   **Physical Inactivity (%):** Percentage of adults (18+) reporting no leisure-time physical activity (age-adjusted).

-   **(FEI) Food Environment Index (1-10):** Composite score indicating community access to healthy food and income levels.

-   **Limited Access to Healthy Food (%):** Percentage of the population that is low income and does not live close to a grocery store

For missing data, we imputed missing predictor values using the state average.
Counties missing data for the response variable (adults with obesity) were excluded from the analysis (a total of 7 counties), resulting in a complete case dataset.

```{r setup}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center",
               echo=FALSE, tidy=TRUE)
```

```{r cleaning}
library(readxl)
library(tidyverse)
theme_set(theme_bw())

data_2025 <- read_excel("C:/Users/helen/OneDrive/Documents/GitHub/obesity-capstone/Peng/2025_data.xlsx", sheet = "Additional Measure Data", skip = 1)

food_index_2025 <- read_excel("C:/Users/helen/OneDrive/Documents/GitHub/obesity-capstone/Peng/2025_data.xlsx", sheet = "Select Measure Data", skip = 1)

data_2025 <- data_2025 |> 
  janitor::clean_names() 

food_index_2025 <- food_index_2025 |> 
  janitor::clean_names() 

data_2025 <- merge(data_2025, food_index_2025, by = "fips") 
data_2025 <- data_2025[, !grepl("percent_ci", names(data_2025))]
data_2025 <- data_2025[, !grepl("national_z", names(data_2025))]
data_2025 <- data_2025[, !grepl("injury_death", names(data_2025))]
data_2025 <- data_2025[, !grepl("percentile", names(data_2025))]
data_2025 <- data_2025[, !grepl("_white", names(data_2025))]
data_2025 <- data_2025[, !grepl("_hispanic", names(data_2025))]
data_2025 <- data_2025[, !grepl("_native", names(data_2025))]
data_2025 <- data_2025[, !grepl("_aian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_asian", names(data_2025))]
data_2025 <- data_2025[, !grepl("_black", names(data_2025))]
data_2025 <- data_2025[, !grepl("dentist", names(data_2025))]
data_2025 <- data_2025[, !grepl("unreliable", names(data_2025))]
data_2025 <- data_2025[, !grepl("deaths", names(data_2025))]

drop_cols <- c("child_mortality_rate","hiv_prevalence_rate", " suicide_rate_age_adjusted", "crude_rate","teen_birth_rate","chlamydia_rate","number_chlamydia_cases", "drug_overdose_mortality_rate","other_primary_care_provider_rate", "other_primary_care_provider_ratio","percent_census_participation", "average_grade_performance_252", " average_grade_performance_246","cohort_size","gender_pay_gap","segregation_index_287","homicide_rate","motor_vehicle_mortality_rate", "percent_disconnected_youth","firearm_fatalities_rate","percent_lacking_support", "percent_low_birth_weight","number_primary_care_physicians","primary_care_physicians_ratio","primary_care_physicians_rate","number_mental_health_providers", "mental_health_provider_rate", "mental_health_provider_ratio", "visits_per_service_area_population", "presence_of_water_violation","average_daily_pm2_5", "state.y","county.y", "percent_with_annual_mammogram", "preventable_hospitalization_rate", "years_of_potential_life_lost_rate","number_firearm_fatalities","child_care_centers_per_1_000_children", "number_firearm_fatalities", "number_child_care_centers","spending_per_pupil","segregation_index_258","high_school_graduation_rate","average_grade_performance_246","disaster_declarations","weeks_in_moderate_or_greater_drought", "life_expectancy", "severe_housing_cost_burden", "percent_adults_with_diabetes","suicide_rate_age_adjusted","percent_fair_or_poor_health", "percent_voter_turnout", "percent_with_disability", "age_adjusted_death_rate", "percent_drive_alone_to_work", "income_ratio", "mens_median_earnings", "womens_median_earnings", "infant_mortality_rate","percent_uninsured_children","percent_uninsured_adults","percent_with_access_to_exercise_opportunities")

data_2025 <- data_2025[,!(names(data_2025) %in% drop_cols)]

data_2025 <- data_2025  |> 
  rename(
    county = county.x,
    state = state.x
  )

old_ct_fips <- c("09001", "09003", "09005", "09007", "09009", "09011", "09013", "09015")  

statewide_food_index <- data_2025 |> 
  filter(is.na(county) & !is.na(food_environment_index)) |> 
  dplyr::select(state, food_environment_index) |> 
  rename(statewide_food_environment_index = food_environment_index)

data_2025 <- data_2025 |> 
  filter(!(fips %in% old_ct_fips)) |> 
  left_join(statewide_food_index, by = "state") |> 
  mutate(
    food_environment_index = if_else(
      is.na(food_environment_index) & !is.na(county),
      statewide_food_environment_index,
      food_environment_index),
      days_above_90f = if_else(days_above_90f > 365, 365, days_above_90f)) |> 
  dplyr::select(-statewide_food_environment_index) |> 
  filter(!is.na(percent_adults_with_obesity))
```

### Exploratory Data Analysis

```{r scatterplot}
#| fig-width: 5
#| fig-height: 3.5
#| fig-cap: "Figure 1: Scatter plots illustrating a positive association between physical inactivity and obesity rates, and a negative association between healthy food access and obesity rates."

library(patchwork)
physical_obesity_plot <- data_2025 |>
  ggplot(aes(x = percent_physically_inactive, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "% Physically Inactive",
    y = "% Adults with Obesity"
  ) 

food_obesity_plot <- data_2025 |>
  ggplot(aes(x = food_environment_index, y = percent_adults_with_obesity)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "gam", se = FALSE) +
  labs(
    x = "Food Environment Index",
    y = "% Adults with Obesity"
  ) 

physical_obesity_plot + food_obesity_plot
```

```{r obesity_map}
#| fig-width: 5
#| fig-height: 3.5
#| fig-cap: "Figure 2: Choropleth map showing substantial regional variation in adult obesity rates across U.S. counties in 2025, with higher prevalence observed in the Midwest and Southeast"

library(usmap)
library(viridis)
library(ggplot2)

min_val <- floor(min(data_2025$percent_adults_with_obesity, na.rm = TRUE))
max_val <- ceiling(max(data_2025$percent_adults_with_obesity, na.rm = TRUE))

custom_breaks <- seq(min_val, max_val, length.out = 5)

plot_usmap(data = data_2025, values = "percent_adults_with_obesity", color = NA) +
  scale_fill_viridis_c(
    option = "inferno",
    direction = -1,
    limits = c(min_val, max_val),
    breaks = custom_breaks,
    labels = round(custom_breaks, 1),
    name = "Obesity Rate (%)",
    na.value = "grey90"
  ) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "bold")
  )
```

## Methods

Based on our exploratory data analysis, we found that the Food Environment Index (FEI) was a more comprehensive predictor than the percentage of limited access to healthy food, and thus selected it for modeling.
Our response variable, adult obesity rate, was originally expressed as a percentage.
We first re-scaled it to a proportion between 0 and 1, denoted as $p$, and then applied the logit transformation:

$$
\text{logit}(p) = \log\left(\frac{p}{1 - p}\right)
$$

This transformation ensured the outcome variable was unbounded, which better satisfies the linearity assumption of regression models.

We initially fit a multiple linear regression model:

$$
\text{logit}(\text{Obesity Rate}) = \beta_0 + \beta_1 \cdot \text{Physical Inactivity} + \beta_2 \cdot \text{Food Environment Index} + \varepsilon
$$

A linear regression model assumes linearity, constant variance (homoscedasticity), independence of observations, and normally distributed residuals.
However, our diagnostic analysis revealed violations of these assumptions, notably due to about 200 influential outliers identified, using the standard Cook’s Distance threshold ($D_i > \frac{4}{n}$), which would negatively affect the model fit.
To address this, we refit the model using Huber regression, a robust technique that reduces the influence of outliers.

To further evaluate the predictive utility of our selected variables, we refit the same model using limited access to healthy food in place of the Food Environment Index.
This parallel model allowed us to directly compare the two predictors and reinforce the conclusion that FEI may offer a stronger and more comprehensive explanation of obesity variation across counties.

## Results

Describe your results.
This can include tables and plots showing your results, as well as text describing how your models worked and the appropriate interpretations of the relevant output.
(Note: Don’t just write out the textbook interpretations of all model coefficients. Instead, interpret the output that is relevant for your question of interest that is framed in the introduction)

```{r}

```

### Model Evaluation

```{r}

```

## Recommendations

**Prioritize Physical Activity Infrastructure**

Investing in infrastructure that promotes physical activity is essential for reducing obesity rates, particularly in counties with high levels of physical inactivity.
Creating walkable communities by expanding sidewalks, bike lanes, and trails makes active transportation more accessible and safe.
A notable example is the *Minneapolis Complete Streets Policy*, which prioritizes pedestrians, bicyclists, and transit users in street design.
This initiative has led to increased biking and walking while supporting safer and more equitable access to physical activity.
Similar models could be adopted in high-obesity counties across the Midwest and South, where urban sprawl often limits pedestrian infrastructure.
Additionally, increasing investment in *Safe Routes to School* programs can further support active lifestyles by improving crosswalks, signage, and sidewalk conditions near schools, helping children build habits of daily activity.

**Improve Food Access with Targeted Interventions**

Food access remains a major barrier in many rural and low-income urban counties, particularly across the South and Midwest.
Mobile food solutions, such as *Veggie Van* in North Carolina and *Fresh Moves Mobile Market* in Chicago, have proven successful in delivering fresh produce to communities with limited access to grocery stores.
These services often pair produce sales with cooking demonstrations or nutrition workshops, making healthy eating more accessible and culturally relevant.
Pop-up farmers markets held at schools, churches, and community centers can further bridge food gaps, especially when coordinated with local public health departments or nonprofit organizations.
These strategies are flexible, community-driven, and can be scaled or adapted based on local needs and food availability.

**Redirect Funding to Hyper-Local Initiatives**

While broad indicators like the Food Environment Index (FEI) are useful for identifying trends, targeted, hyper-local strategies often yield greater impact.
Redirecting funding from large-scale programs to smaller, community-embedded efforts can better address the specific needs of a neighborhood or county.
For example, in Louisville, Kentucky, the *Fresh Stop Markets* program partners with local farmers to provide subsidized CSA (Community Supported Agriculture) boxes to low-income residents, improving access to fresh produce while supporting regional agriculture.
Such programs offer more tailored support than generic interventions, allowing communities to engage directly in both the design and implementation of solutions.
This approach ensures that resources are used where they are most needed, with maximum relevance and efficiency.

## Discussion

### Key Takeaways

This study finds that physical inactivity is the strongest and most consistent predictor of adult obesity at the county level.
While improving access to healthy food is often seen as a key intervention, the data suggest that access alone is insufficient—particularly if physical inactivity prevents individuals from fully benefiting from available food resources.
Notably, the Food Environment Index (FEI) captures more variability in obesity rates than any single variable.
By combining measures of both food access and economic constraints, the FEI provides a more holistic representation of environmental and socioeconomic barriers to health.
These findings highlight the importance of addressing physical activity and food access together, rather than treating them as isolated factors.

### Future Work

To build on these findings, several avenues for future research are recommended.
First, incorporating data on the number of parks, wellness centers, and recreational facilities per county—and mapping their distribution—can help clarify the role of environmental supports for physical activity.
Additionally, analyzing temporal trends in fitness facility usage may offer insights into behavioral patterns and access disparities.
Future studies should also explore other relevant variables that could influence obesity, such as public transportation availability, climate, and healthcare access.
Finally, applying causal inference methods, such as instrumental variable analysis or matching methods, could help determine whether physical inactivity directly contributes to obesity, as current results are based on observed correlations and cannot confirm causation.
